---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  dev = "ragg_png",
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# gridmicrotex

<!-- badges: start -->
[![R-CMD-check](https://github.com/adayim/gridmicrotex/workflows/R-CMD-check/badge.svg)](https://github.com/adayim/gridmicrotex/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/gridmicrotex)](https://CRAN.R-project.org/package=gridmicrotex)
[![CRAN download](https://cranlogs.r-pkg.org/badges/grand-total/gridmicrotex)](https://cran.r-project.org/package=gridmicrotex)
[![codecov](https://codecov.io/gh/adayim/gridmicrotex/branch/main/graph/badge.svg?token=mzvaYDMPNc)](https://app.codecov.io/gh/adayim/gridmicrotex)
<!-- badges: end -->

Render LaTeX math expressions --- and markdown containing them --- as native
R **grid** graphics objects, with no external LaTeX installation required.

gridmicrotex embeds the [MicroTeX](https://github.com/NanoMichael/MicroTeX)
C++ layout engine to parse LaTeX, compute the full box model, and produce
resolution-independent vector output (paths, lines, rectangles) that works on
any R graphics device.

## Disclaimer
**A note on development**: This package was developed as a proof of concept for AI-assisted package creation. I designed the architecture and specification, and the core C++ integration (via [MicroTeX](https://github.com/NanoMichael/MicroTeX)) was largely facilitated by AI, with my review and oversight of the design and final outputs. I am sharing it because it works, and I hope that others will find it useful. Contributions, bug reports and improvements from the community are very welcome.


## Installation

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("adayim/gridmicrotex")
```

## Examples

```{r example-basic, fig.height=0.8, fig.width=3, out.width="50%"}
library(gridmicrotex)
library(grid)

grid.newpage()
grid.latex("x = \\frac{\\textcolor{red}{-b} \\pm \\sqrt{b^{2} - 4ac}}{2a}", 
           gp = grid::gpar(fontsize = 30))
```


By default, the input is treated as LaTeX math mode ("mixed" mode), which wraps non-math text in `\text{}` and preserves math expressions as-is. Use `$...$` or `\\(...\\)` delimiters to render math. The `"x = "` in the equation above treated as text. Use `input_mode = "math"` to treat the whole string as math mode and render text with `\\text{}`. You can change this with global option `latex_options(input_mode = "math")`.


### Composing with other grobs

The grob can be placed alongside other grid objects:

```{r example-compose, fig.height=0.8, fig.width=2, out.width="30%"}
latex_options(input_mode = "math")
g <- latex_grob("\\frac{a}{b}", gp = grid::gpar(fontsize = 30))
grid.newpage()
# A blue box behind the formula
grid.rect(
  x = 0.5, y = 0.5,
  width = grobWidth(g) + unit(10, "bigpts"),
  height = grobHeight(g) + unit(10, "bigpts"),
  gp = gpar(fill = "#e8f0fe", col = "#4285f4", lwd = 2)
)

# The formula itself
grid.draw(g)
```

### Multiple expressions

```{r example-multiple, fig.height=3, fig.width=4, out.width="40%"}
exprs <- c(
  "E = mc^{2}",
  "e^{i\\pi} + 1 = 0",
  "\\nabla \\times \\vec{E} = -\\frac{\\partial \\vec{B}}{\\partial t}",
  "\\binom{n}{k} = \\frac{n!}{k!(n-k)!}"
)

grid.newpage()
for (i in seq_along(exprs)) {
  grid.latex(
    exprs[i],
    x = unit(0.5, "npc"),
    y = unit(1 - i / (length(exprs) + 1), "npc"),
    gp = grid::gpar(fontsize = 28)
  )
}
```

### Mixed text and math

You can use `r"()"` raw strings to write LaTeX with regular newlines and quotes without escaping. Use `\text{}` to embed regular text within math expressions:


```{r example-mixed-definition, fig.height=1.2, fig.width=6, out.width="50%"}
grid.newpage()
grid.latex(
  r"(f(x) = \begin{cases} x^2 & \text{if } x \geq 0 \\ -x & \text{otherwise} \end{cases})",
  gp = grid::gpar(fontsize = 26)
)
```

## Markdown

`grid.markdown()` renders markdown, with `$math$` inline. Unlike the other
markdown-in-grid packages, the maths is real LaTeX rather than plotmath:

```{r example-markdown, fig.height=0.8, fig.width=6, out.width="70%"}
grid.newpage()
grid.markdown(
  "The **fitted** slope is $\\hat{\\beta}_1 = 0.42$ (*p* < 0.001).",
  x = 0.02, hjust = 0, gp = gpar(fontsize = 20)
)
```

Colour, font and size come from inline HTML, as they must: markdown itself
defines no syntax for them.

```{r example-markdown-html, fig.height=0.8, fig.width=6, out.width="70%"}
grid.newpage()
grid.markdown(
  paste0('Set in <span style="font-family:serif">serif</span>, ',
         'H<sub>0</sub> <u>rejected</u> at ',
         '<span style="color:#B22222">5%</span>.'),
  x = 0.02, hjust = 0, gp = gpar(fontsize = 20)
)
```

`markdown_box_grob()` lays out a whole document --- headings, lists, quotes,
code, tables and images --- as a stack of grobs. See
`vignette("markdown")`.


## ggplot2 integration

Use `geom_latex()` to place LaTeX labels at data coordinates, and
`element_latex()` for LaTeX-rendered axis titles:

```{r example-ggplot2-geom, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, out.width="80%"}
library(ggplot2)
# Add a LaTeX table as an annotation
tab_str <- r"(\begin{tabular}{c|c} \text{A} & B^2 \\ \hline 1 & \cellcolor{#00bde5}2 \\ 3 & 4 \end{tabular})"

df <- data.frame(x = 1:3, y = 1:3,
                 eq = c("x^2", "\\frac{a}{b}", "\\sum_{i=1}^n x_i"))
ggplot(df, aes(x, y, label = eq)) + 
  geom_latex() +
  annotate("latex", x = 1, y = 2.7, label = tab_str, size = 12) +
  labs(x = "$\\beta_1 \\cdot x + \\beta_0$") +
  theme(axis.title.x = element_latex())
```


ggplot2 is a soft dependency --- the core functions work without it.
See `vignette("ggplot2-integration")` for more examples.


## Comparison

| Approach       | LaTeX required? | Device independent? | Math coverage | Markdown |
|:---------------|:---------------:|:-------------------:|:-------------:|:--------:|
| `tikzDevice`   | Yes             | No                  | Full          | No       |
| `xdvir`        | Yes             | No                  | Full          | No       |
| `latexpdf`     | Yes             | No                  | Full (tables) | No       |
| `latex2exp`    | No              | Yes                 | Limited       | No       |
| `plotmath`     | No              | Yes                 | Limited       | No       |
| `gridtext`     | No              | Yes                 | None          | Yes      |
| `marquee`      | No              | Yes                 | None          | Yes      |
| **gridmicrotex** | **No**       | **Yes**             | **Broad**     | **Yes**  |


## How it works

1. Your LaTeX string is parsed by MicroTeX's C++ engine into a TeX box model
2. A custom `Graphics2D` recorder captures every draw operation (glyph paths,
   lines, rectangles) with exact coordinates
3. The layout crosses the C++/R boundary as a data frame
4. R converts each record into native grid primitives (`pathGrob`,
   `segmentsGrob`, `rectGrob`)
5. The result is a `gTree` that renders on any device at any resolution

By default, math glyphs are rendered in typeface mode as native text
using the selected math font, which keeps PDF/SVG output selectable and
searchable on devices with font embedding support.

When `render_mode = "path"` is used (or when automatic fallback is
triggered on unsupported devices), glyphs are drawn as filled vector
paths for consistent rendering everywhere.

Make sure to use `ragg::agg_png()`, `svglite::svglite()` or `grDevices::cairo_pdf()` for best results, as some older devices may not support the full range of path operations.

## Graphics backend

The default graphics device on Windows (`windows()`) and macOS
(`quartz()`) may not find the bundled math fonts, producing warnings
like:

```
font family not found in Windows font database
```

To avoid this, switch to a modern graphics backend that uses
[systemfonts](https://CRAN.R-project.org/package=systemfonts) for font
resolution:

```r
# For knitr / R Markdown — add to your setup chunk:
knitr::opts_chunk$set(dev = "ragg_png")

# For interactive use:
options(device = function(...) ragg::agg_png(tempfile(fileext = ".png"), ...))
```

Recommended backends:

| Backend               | Format | Package               |
|:----------------------|:-------|:----------------------|
| `ragg::agg_png()`     | PNG    | [ragg](https://CRAN.R-project.org/package=ragg)       |
| `svglite::svglite()`  | SVG    | [svglite](https://CRAN.R-project.org/package=svglite) |
| `grDevices::cairo_pdf()` | PDF | Base R (Cairo build)  |

Alternatively, use `render_mode = "path"` to bypass font lookup
entirely — glyphs are drawn as vector paths, which works on all
devices but produces non-selectable text in PDF/SVG.

