---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  dev = "ragg_png",
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# gridmicrotex

<!-- badges: start -->
[![R-CMD-check](https://github.com/adayim/gridmicrotex/workflows/R-CMD-check/badge.svg)](https://github.com/adayim/gridmicrotex/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/gridmicrotex)](https://CRAN.R-project.org/package=gridmicrotex)
[![codecov](https://codecov.io/gh/adayim/gridmicrotex/branch/main/graph/badge.svg?token=mzvaYDMPNc)](https://app.codecov.io/gh/adayim/gridmicrotex)
<!-- badges: end -->

Render LaTeX math expressions as native R **grid** graphics objects, with no
external LaTeX installation required.

gridmicrotex embeds the [MicroTeX](https://github.com/NanoMichael/MicroTeX)
C++ layout engine to parse LaTeX, compute the full box model, and produce
resolution-independent vector output (paths, lines, rectangles) that works on
any R graphics device.

## Disclaimer
**A note on development**: This package was developed as a proof of concept exploring AI-assisted package creation. The architecture and specification were designed by me, and the the core C++ integration (via [MicroTeX](https://github.com/NanoMichael/MicroTeX)) was facilitated largely by AI, with my review and oversight of the final output. I'm sharing it because it works and I hope others find it useful. Contributions, bug reports, and improvements from the community are very welcome.


## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("adayim/gridmicrotex")
```

## Examples

```{r example-basic, fig.height=1.5, fig.width=4}
library(gridmicrotex)
library(grid)

grid.newpage()
grid.latex("x = \\frac{\\textcolor{red}{-b} \\pm \\sqrt{b^{2} - 4ac}}{2a}", fontsize = 30)
```

### Composing with other grobs

The grob can be placed alongside other grid objects:

```{r example-compose, fig.height=2.5, fig.width=6}
g <- latex_grob("\\frac{a}{b}", fontsize = 30)
grid.newpage()
# A blue box behind the formula
grid.rect(
  x = 0.5, y = 0.5,
  width = grobWidth(g) + unit(10, "bigpts"),
  height = grobHeight(g) + unit(10, "bigpts"),
  gp = gpar(fill = "#e8f0fe", col = "#4285f4", lwd = 2)
)

# The formula itself
grid.draw(g)
```

### Multiple expressions

```{r example-multiple, fig.height=4, fig.width=6}
exprs <- c(
  "E = mc^{2}",
  "e^{i\\pi} + 1 = 0",
  "\\nabla \\times \\vec{E} = -\\frac{\\partial \\vec{B}}{\\partial t}",
  "\\binom{n}{k} = \\frac{n!}{k!(n-k)!}"
)

grid.newpage()
for (i in seq_along(exprs)) {
  grid.latex(
    exprs[i],
    x = unit(0.5, "npc"),
    y = unit(1 - i / (length(exprs) + 1), "npc"),
    fontsize = 28
  )
}
```

### Mixed text and math

Use `\text{}` to embed regular text within math expressions:


```{r example-mixed-definition, fig.height=2.5, fig.width=8}
grid.newpage()
grid.latex(
  "f(x) = \\begin{cases} x^2 & \\text{if } x \\geq 0 \\\\ -x & \\text{otherwise} \\end{cases}",
  fontsize = 26
)
```

## CJK and multilingual text

Non-math text via `\text{}` supports CJK and other scripts. Font settings from
`gp` (`fontfamily`, `fontface`) apply to text only --- math rendering always
uses the selected math font:


```{r example-mixed-cjk-math, fig.height=1.5, fig.width=8}
grid.newpage()
grid.latex("\\text{如果 } x > 0 \\text{ 则 } y = x^2", fontsize = 24,
           gp = gpar(fontfamily = "sans"))
```

## ggplot2 integration

Use `geom_latex()` to place LaTeX labels at data coordinates, and
`element_latex()` for LaTeX-rendered axis titles:

```{r example-ggplot2-geom, fig.height=4, fig.width=6}
library(ggplot2)

df <- data.frame(x = 1:3, y = 1:3,
                 eq = c("x^2", "\\frac{a}{b}", "\\sum_{i=1}^n x_i"))
ggplot(df, aes(x, y, label = eq)) + 
  geom_latex() +
  labs(x = "$\\beta_1 \\cdot x + \\beta_0$") +
  theme(axis.title.x = element_latex())
```


ggplot2 is a soft dependency --- the core functions work without it.
See `vignette("ggplot2-integration")` for more examples.

## Features

- Full LaTeX math: fractions, roots, integrals, matrices, Greek, accents,
  extensible delimiters
- Two bundled math fonts: Latin Modern Math, XITS Math
- Color support: `\textcolor{}`, `\color{}`
- Font variants: `\mathbb{}`, `\mathcal{}`, `\mathfrak{}`
- CJK/multilingual text in `\text{}`
- ggplot2 geom, annotation, and theme element (S7-compatible with ggplot2 >= 4.0)
- Resolution-independent vector output on all R devices
- No external LaTeX installation required

## Comparison

| Approach       | LaTeX required? | Device independent? | Vector? | Math coverage |
|:---------------|:---------------:|:-------------------:|:-------:|:-------------:|
| `tikzDevice`   | Yes             | No                  | Yes     | Full          |
| `xdvir`        | Yes             | No                  | Yes     | Full          |
| `latexpdf`     | Yes             | No                  | Yes     | Full (tables) |
| `latex2exp`    | No              | Yes                 | Yes     | Limited       |
| `plotmath`     | No              | Yes                 | Yes     | Limited       |
| **gridmicrotex** | **No**       | **Yes**             | **Yes** | **Broad**     |


## How it works

1. Your LaTeX string is parsed by MicroTeX's C++ engine into a TeX box model
2. A custom `Graphics2D` recorder captures every draw operation (glyph paths,
   lines, rectangles) with exact coordinates
3. The layout crosses the C++/R boundary as a data frame
4. R converts each record into native grid primitives (`pathGrob`,
   `segmentsGrob`, `rectGrob`)
5. The result is a `gTree` that renders on any device at any resolution

All math glyphs are rendered as filled vector paths from the bundled
Latin Modern Math font, so output is crisp at any size and fully
resolution-independent.

Make sure to use `ragg::agg_png()`, `svglite::svglite()` or `grDevices::cairo_pdf()` for best results, as some older devices may not support the full range of path operations.

## Graphics backend

The default graphics device on Windows (`windows()`) and macOS
(`quartz()`) may not find the bundled math fonts, producing warnings
like:

```
font family not found in Windows font database
```

To avoid this, switch to a modern graphics backend that uses
[systemfonts](https://CRAN.R-project.org/package=systemfonts) for font
resolution:

```r
# For knitr / R Markdown — add to your setup chunk:
knitr::opts_chunk$set(dev = "ragg_png")

# For interactive use:
options(device = function(...) ragg::agg_png(tempfile(fileext = ".png"), ...))
```

Recommended backends:

| Backend               | Format | Package               |
|:----------------------|:-------|:----------------------|
| `ragg::agg_png()`     | PNG    | [ragg](https://CRAN.R-project.org/package=ragg)       |
| `svglite::svglite()`  | SVG    | [svglite](https://CRAN.R-project.org/package=svglite) |
| `grDevices::cairo_pdf()` | PDF | Base R (Cairo build)  |

Alternatively, use `render_mode = "path"` to bypass font lookup
entirely — glyphs are drawn as vector paths, which works on all
devices but produces non-selectable text in PDF/SVG.

