---
title: "LaTeX math in base R graphics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LaTeX math in base R graphics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "png"
)
```

Turn on `latex_options(device_math = TRUE)` and any plot label written
with `$...$` is typeset as LaTeX. The rest of your plotting code stays
the same.

## A first plot

```{r first, fig.height=3.8, fig.width=6, out.width="75%"}
library(gridmicrotex)
library(grid)

latex_options(device_math = TRUE)

plot(1:10, (1:10)^2,
     main = r"(Slope $\hat{\beta}_1 = \sum_{i=1}^{n} x_i^2$)",
     xlab = "Cost $5-$10 per unit",
     ylab = r"($\frac{y}{2}$)")
text(3, 80, r"($\int_0^\infty e^{-x^2}\,dx$)", col = "steelblue")
```

The title, y-axis label and annotation are typeset. The x-axis label has
no math in it, so it is drawn as written.

In R Markdown, set the option in the setup chunk. If you turn it off,
do so in a later chunk than the plot, or the plot shows the literal
`$...$`.

## More than plotmath

Tables, display-style operators and other LaTeX work in any label. The
coefficient table below is drawn by an ordinary `text()` call:

```{r beyond, fig.height=4.6, fig.width=6.5}
set.seed(42)
x   <- seq(0, 3, length.out = 60)
y   <- 2 + 1.4 * x^2 + rnorm(60, sd = 1.1)
fit <- lm(y ~ poly(x, 2, raw = TRUE))
cf  <- signif(coef(summary(fit)), 3)

title_eq <- r"($$\hat{y} = \beta_0 + \beta_1 x + \beta_2 x^2$$)"

# Make room for the tall title
h   <- latex_dims(title_eq, gp = gpar(fontsize = par("ps")))$height
top <- ceiling(convertHeight(h, "bigpts", TRUE) /
               (par("cin")[2] * 72 * par("mex"))) + 2

par(mar = c(4.5, 5.5, top, 2))
plot(x, y, pch = 19, col = "grey45", cex = 0.7, ylim = c(0, 22),
     main = title_eq, xlab = r"($x$)",
     ylab = r"($y = f(x) + \varepsilon$)")
lines(x, predict(fit), col = "#B22222", lwd = 2)

# Build the table as a single line of text
tab <- sprintf(paste0(
  r"(\begin{array}{|l|r|r|}\hline)",
  r"(\rowcolor{#E8F0FE}\text{Term} & \hat{\beta} & \text{SE}\\\hline)",
  r"(\beta_0 & %s & %s\\)",
  r"(\beta_1 & %s & %s\\)",
  r"(\beta_2 & %s & %s\\\hline\end{array})"),
  cf[1, 1], cf[1, 2], cf[2, 1], cf[2, 2], cf[3, 1], cf[3, 2])

text(0.05, 18, tab, adj = c(0, 0.5), cex = 0.75)
text(2.95, 1.2, r"($\varepsilon \sim \mathcal{N}(0, \sigma^2)$)",
     adj = c(1, 0), cex = 0.95, col = "#B22222")
```

Two tips from this example:

- Keep a formula on one line. A label containing a newline is not
  recognised as math.
- Anchor a tall label by its centre (`adj = c(0, 0.5)`) and leave room
  around it; see [Tall formulas](#tall-formulas).

## Other packages

The option works for any plot drawn on an R graphics device: base plot
methods such as `plot(lm)`, `hist()` and `legend()`, lattice, grid, and
ggplot2 (axis titles, `geom_text()`, facet strips, legend keys). It does
not reach plotly or rgl's `text3d()` labels.

For ggplot2, `geom_latex()` and `element_latex()` are often the better
choice; see `vignette("ggplot2-integration")`.

## When is a label treated as math?

A label is typeset only when every `$` in it is closed and each `$...$`
pair holds something that looks like math. Everyday dollar signs are
left alone:

| Label | Rendered as |
|:--|:--|
| `"Revenue ($)"` | text |
| `"Price is $5 today"` | text |
| `"Cost $5-$10"` | text |
| `"Budget $1,000 to $5,000 with $x$ shown"` | text |
| `"Histogram of df$a_b - df$c_d"` | text |
| `"$x$"` | math |
| `"$x^2$"` | math |
| `"$y = 2x + 1$"` | math |
| `"Slope $\\hat{\\beta}_1$"` | math |
| `"\\(\\alpha\\)"` | math |

`$$...$$`, `\(...\)` and `\[...\]` also work. Write `\$` for a literal
dollar sign in a label that also has math.

## Tall formulas {#tall-formulas}

Widths are measured correctly, but heights are not: R sizes margins,
titles and `legend()` boxes for one line of text, so a tall formula can
overflow them. Measure it with `latex_dims()` and widen the margin:

```{r margin}
h  <- latex_dims(r"($\frac{a}{b}$)", gp = gpar(fontsize = par("ps")))$height
bp <- convertHeight(h, "bigpts", TRUE)

# Lines of margin needed
bp / (par("cin")[2] * 72 * par("mex"))
```

Round that up and add it to the matching side of `par(mar = )`.

## Limitations

- The option applies to every plot in the session, including plots from
  other packages. Labels without math are not changed.
- Math is drawn as outlines, so it cannot be selected in a PDF or SVG.
- Bold comes from `\textbf{}`, not from the plot's font face.
- `\includegraphics` is skipped, and rounded box corners are square.
- A label that cannot be read is drawn as plain text, without a warning.
- It does not work together with `showtext::showtext_auto()`.
- `expression()` labels are not affected.
- corrplot reads a label starting with `$` itself; start it with
  something else.

```{r off, include = FALSE}
latex_options(device_math = FALSE)
```
