---
title: "LaTeX math in base R graphics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LaTeX math in base R graphics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "png"
)
```

Base graphics is the plotting system R starts with, and its only route to
mathematical notation has been plotmath: `expression()` labels with their
own small syntax and no tables, alignment or real LaTeX. gridmicrotex adds
one switch that changes this. Turn on `latex_options(device_math = TRUE)`
and any label written with `$...$` is typeset by MicroTeX, the same engine
behind `latex_grob()`, with no other change to your plotting code.

## A first plot

```{r first, fig.height=3.8, fig.width=6, out.width="75%"}
library(gridmicrotex)
library(grid)   # gpar() and unit conversion, used further down

latex_options(device_math = TRUE)

plot(1:10, (1:10)^2,
     main = r"(Slope $\hat{\beta}_1 = \sum_{i=1}^{n} x_i^2$)",
     xlab = "Cost $5-$10 per unit",
     ylab = r"($\frac{y}{2}$)")
text(3, 80, r"($\int_0^\infty e^{-x^2}\,dx$)", col = "steelblue")
```

`plot()` and `text()` are the ordinary base functions. The title, the
y-axis title and the annotation are typeset, while `"Cost $5-$10 per unit"`
is drawn exactly as written, because a label is only treated as math when
it clearly contains some (see
[When is a label treated as math?](#when-is-a-label-treated-as-math)).
The labels are written as raw strings, `r"(...)"`, so their backslashes
need no doubling.

The option stays on for the rest of this vignette and is switched off at
the end.

## In R Markdown

Set the option once, in your setup chunk. knitr captures a chunk's plots
*after* its last line has run, so switching the option off at the end of
the chunk that draws leaves the figure unrendered, with the labels coming
out as literal `$...$`. If you need to turn it off, do it in a later chunk.

## Beyond plotmath

This is the whole LaTeX engine, not a translation layer, so a base plot
can carry things plotmath has no syntax for at all: a ruled table with a
coloured header row, display-style operators, calligraphic and blackboard
alphabets, automatic delimiter sizing. The coefficient table below is a
`\begin{array}` with `\hline` rules and `\rowcolor`, drawn by an ordinary
`text()` call.

```{r beyond, fig.height=4.6, fig.width=6.5}
set.seed(42)
x   <- seq(0, 3, length.out = 60)
y   <- 2 + 1.4 * x^2 + rnorm(60, sd = 1.1)
fit <- lm(y ~ poly(x, 2, raw = TRUE))
cf  <- signif(coef(summary(fit)), 3)

title_eq <- r"($$\hat{y} = \beta_0 + \beta_1 x + \beta_2 x^2$$)"

# A display-style title is taller than one text line, so take the top
# margin from the measured height instead of guessing at it.
h   <- latex_dims(title_eq, gp = gpar(fontsize = par("ps")))$height
top <- ceiling(convertHeight(h, "bigpts", TRUE) /
               (par("cin")[2] * 72 * par("mex"))) + 2

par(mar = c(4.5, 5.5, top, 2))
plot(x, y, pch = 19, col = "grey45", cex = 0.7, ylim = c(0, 22),
     main = title_eq, xlab = r"($x$)",
     ylab = r"($y = f(x) + \varepsilon$)")
lines(x, predict(fit), col = "#B22222", lwd = 2)

# Keep a formula on one physical line: R hands the device a multi-line
# label one line at a time, so a table split across source lines would
# never reach it whole.
tab <- sprintf(paste0(
  r"(\begin{array}{|l|r|r|}\hline)",
  r"(\rowcolor{#E8F0FE}\text{Term} & \hat{\beta} & \text{SE}\\\hline)",
  r"(\beta_0 & %s & %s\\)",
  r"(\beta_1 & %s & %s\\)",
  r"(\beta_2 & %s & %s\\\hline\end{array})"),
  cf[1, 1], cf[1, 2], cf[2, 1], cf[2, 2], cf[3, 1], cf[3, 2])

text(0.05, 18, tab, adj = c(0, 0.5), cex = 0.75)
text(2.95, 1.2, r"($\varepsilon \sim \mathcal{N}(0, \sigma^2)$)",
     adj = c(1, 0), cex = 0.95, col = "#B22222")
```

Two placement notes, both consequences of text heights, which cannot be
corrected (see
[Reserving room for a tall formula](#reserving-room-for-a-tall-formula)).
The table is anchored with `adj = c(0, 0.5)` and given headroom through
`ylim`, because R positions a top-anchored label using the *font* ascent:
a four-row table anchored at the very top of the panel would have its
header clipped away. And a formula written across several source lines
arrives at the device one line at a time and so is never recognised, which
is why `tab` is assembled with `paste0()` into a single line.

## How it works

`latex_grob()` asks MicroTeX to lay a formula out and turns the result into
grid grobs. `device_math` uses the same layout but skips the grobs:
gridmicrotex registers with R's graphics engine and, on every device it
opens, takes over the two callbacks through which all text is measured and
drawn. A string with no `$`, `\(`, `\[` or `\begin{` in it is handed
straight back to the device. Anything else is checked against the rule
below and, if it is math, laid out by MicroTeX and drawn with the device's
own line, polygon and text primitives.

### Any plot drawn to an R device

Every R graphics system draws its text through those same callbacks, so
the switch reaches far beyond `plot()`. Checked so far: base methods such
as `plot(lm)`, `hist()` and `legend()`, and the plot methods of survival,
igraph, plotrix and gplots; lattice; grid; ggplot2, including figures
assembled with patchwork, gridExtra and cowplot. In ggplot2 that means axis
titles, `geom_text()` and `geom_label()` labels, facet strips and legend
keys all accept `$...$` without a single gridmicrotex function, as
`vignette("ggplot2-integration")` shows.

That lets you keep the code and packages you already use.
`element_latex()` and `geom_latex()` are still worth reaching for in
ggplot2 when a label is tall enough that ggplot2 has to reserve its real
height, or when you want the rendering confined to one layer or theme
element rather than the whole session.

### Output that is not drawn on an R device

plotly and rgl draw somewhere other than an R graphics device, so they were
checked separately.

- **plotly** renders in the browser. `ggplotly()` measures the plot on an R
  device while converting it, but the labels it hands to plotly.js stay
  plain strings, and its output is identical with `device_math` on or off.
  plotly has its own route to math instead: `plotly::config()` takes a
  `mathjax` argument for MathJax rendering in the browser.
- **rgl** draws with OpenGL, and ordinary `text3d()` labels are drawn by
  rgl itself, so they stay literal. The exception is `plotmath3d()`, which
  `text3d(..., usePlotmath = TRUE)` also calls: it draws the label into an R
  `png()` device and shows that image in the scene, so a `$...$` label there
  does become real math. Being an image, it keeps the resolution it was
  drawn at rather than scaling like vector text.

## When is a label treated as math?

A dollar sign is far too common in a real axis label to treat as a
delimiter on sight: `"Cost $5-$10 per unit"` has to survive as plain text.
So a label is intercepted only when **every** delimiter in it is closed,
*and* either the whole label is one formula or **every** `$...$` pair wraps
something that actually looks like math (a command, a script, or a lone
variable). A `$` straight after a name is R's column accessor, never a
delimiter. The delimiters are the ones `latex_wrap()` uses everywhere
else: `$...$`, `$$...$$`, `\(...\)` and `\[...\]`.

| Label | Rendered as | Why |
|:--|:--|:--|
| `"Revenue ($)"` | text | one unclosed `$` |
| `"Price is $5 today"` | text | one unclosed `$` |
| `"Cost $5-$10"` | text | closed, but `5-` is not math |
| `"Budget $1,000 to $5,000 with $x$ shown"` | text | `1,000 to ` is not math |
| `"Histogram of df$a_b - df$c_d"` | text | `$` after a name is column access |
| `"$x$"` | math | a lone variable |
| `"$x^2$"` | math | a script |
| `"$y = 2x + 1$"` | math | the whole label is one formula |
| `"Slope $\\hat{\\beta}_1$"` | math | a command |
| `"\\(\\alpha\\)"` | math | explicit delimiters |

Every pair has to qualify, not merely one of them: `latex_wrap()` cannot be
told to treat one span as math and leave another alone, so a single good
`$x$` would otherwise drag the currency in with it.

Everything that fails the test is handed to the device untouched, byte for
byte. Turning the option on cannot change how an existing plot looks unless
a label really does contain math. Write `\$` for a literal dollar next to
real math in the same label.

## Reserving room for a tall formula

Widths are exact while the option is on: `strwidth()` reports the formula's
real width, so labels centre correctly. **Heights are not, and cannot be.**
R works out text height from the font and never from the string:
`strheight()` returns the same value for `"abc"` as for
`$\sum_{i=1}^{n}\frac{x_i}{2}$`. There is no callback a device could use to
answer "how tall is this string", and the per-character metric callback
that does exist is handed one character code at a time, never the string,
so enlarging it would make *every* line of text on the device taller. A
tall formula can therefore overflow a `legend()` box, the margin
`par("mar")` set aside for it, or a ggplot2 title or facet strip.

Measure it with `latex_dims()` and reserve the room yourself:

```{r margin}
h  <- latex_dims(r"($\frac{a}{b}$)", gp = gpar(fontsize = par("ps")))$height
bp <- convertHeight(h, "bigpts", TRUE)

# par(mar) counts lines of par("cin"), so convert in those units rather
# than grid's "lines": the two agree only while gpar$fontsize happens to
# equal par("ps").
bp / (par("cin")[2] * 72 * par("mex"))
```

Round that up and add it to the relevant side of `par(mar = )`.

## Side effects

`device_math` changes what the graphics device does for the rest of the
session, so it is worth knowing exactly what it reaches.

- **It is global.** Every device opened while it is on is affected,
  including text drawn by packages you did not write. Labels without math
  are passed through byte for byte, so only labels that really contain
  math change.
- **Heights come from the font.** Widths are exact, but margins, `legend()`
  boxes, ggplot2 titles and facet strips are all sized from the font, so a
  tall formula can overflow them.
- **A label split over lines is not recognised.** R hands the device one
  line at a time, so a formula containing a newline never arrives whole.
- **Math is drawn as outlines**, so it is not selectable or searchable in a
  PDF or SVG, and files are larger than with `latex_grob()`.
- **Font face comes from the LaTeX.** Prose in a bold label is drawn plain
  unless it is wrapped in `\textbf{}`.
- **Some constructs are dropped.** `\includegraphics` is left out and
  rounded box corners are drawn square.
- **Failures are quiet.** A label that cannot be laid out is drawn as
  literal text, and warnings raised while laying it out are suppressed:
  there is no good moment to show them in the middle of drawing a plot.
- **showtext wins.** `showtext::showtext_auto()` replaces the same device
  callbacks when a plot starts, so with showtext on the labels come out as
  literal `$...$`. Use one or the other.
- **A package that gives `$` its own meaning acts first.** corrplot, for
  instance, parses a label beginning with `$` as a plotmath expression and
  fails before the device sees it; begin the label with anything else.
- **`expression()` labels are untouched**, because R lays plotmath out
  inside the graphics engine, where a device never sees it.
- **Only R graphics devices are reached.** plotly and other htmlwidgets are
  unaffected, and so are rgl's own `text3d()` labels. rgl's `plotmath3d()`,
  and `text3d(usePlotmath = TRUE)` which calls it, draws through an R device
  and does pick it up.

`?latex_options` has the same list.

```{r off, include = FALSE}
latex_options(device_math = FALSE)
```
