---
title: "Typesetting a document"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Typesetting a document}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "png"
)
library(gridmicrotex)
library(grid)
```

A grid page saved as PDF is a document. With `input_mode = "document"`,
`latex_grob()` and `grid.latex()` read the body of a LaTeX document ---
paragraphs, headings, displayed equations, lists and tables --- by
LaTeX's own rules, so a methods section, a report's appendix or a long
caption can be set in R without a LaTeX installation.

## A document body

This is ordinary LaTeX, as it would sit between `\begin{document}` and
`\end{document}`:

```{r body}
body <- r"(\section{Methods}
We fit a straight line to $n$ observations $(x_i, y_i)$ by least
squares, which chooses the intercept and slope that minimise the sum
of squared residuals
\[ S(\beta_0, \beta_1) = \sum_{i=1}^{n} (y_i - \beta_0 - \beta_1 x_i)^2. \]
Setting both partial derivatives to zero gives the estimates
\begin{align*}
  \hat\beta_1 &= \frac{\sum_i (x_i - \bar x)(y_i - \bar y)}{\sum_i (x_i - \bar x)^2}, \\
  \hat\beta_0 &= \bar y - \hat\beta_1 \bar x.
\end{align*}

The fitted values are $\hat y_i = \hat\beta_0 + \hat\beta_1 x_i$, and
the residuals $e_i = y_i - \hat y_i$ are what is left over.

\subsection{Assumptions}
The errors are taken to be
\begin{itemize}
  \item independent of one another,
  \item of constant variance $\sigma^2$, and
  \item normally distributed --- which matters only for the tests.
\end{itemize}
)"
```

`max_width` is the text width, in big points (1/72 inch). Paragraphs are
broken to it and displays are centred in it:

```{r draw-body, fig.width = 6, fig.height = 5.2, out.width = "90%"}
grid.newpage()
grid.latex(body, input_mode = "document", max_width = 5.6 * 72,
           x = 0.03, y = 0.97, hjust = 0, vjust = 1,
           gp = gpar(fontsize = 11))
```

## How the body is read

As LaTeX reads it, with nothing of our own:

- **Paragraphs.** A line end is a space. A blank line, or `\par`,
  starts a paragraph, whose first line is indented by TeX's `\parindent`
  (1.5 em); there is no space between paragraphs. `\noindent` leaves one
  flush.
- **Headings.** `\section`, `\subsection` and `\subsubsection` are
  numbered 1, 1.1, 1.1.1 and set bold, on a line of their own;
  `\paragraph` runs into its paragraph. A starred heading has no number.
  The paragraph after a heading is not indented.
- **Math.** `$…$` and `\(…\)` are inline. `\[…\]`, `$$…$$` and the
  display environments --- `equation`, `align`, `gather`, `multline`,
  `eqnarray`, starred or not --- are centred on a line of their own, and
  the paragraph goes on after them unindented, as in LaTeX. Equations are
  not numbered.
- **Lists and tables.** `itemize` and `enumerate` are set as blocks;
  their items, like a `tabular`'s cells, are text, with math between
  `$…$`.
- **Text.** `--` and `---` are dashes, `` ` `` and `'` (single or
  doubled) curly quotes, and `~` a space the line is not broken at.
  `\centering` and the `center` environment centre their lines, and a
  `table` or `figure` is set apart where it is written, with its
  `\caption` on a line of its own.

Anything LaTeX would stop on is drawn as well as it can be, and one
warning lists each problem with its line and column in the body.

## The text width

Without `max_width` every paragraph is a single line, however long. With
it, lines are filled one at a time and left ragged; `justify = TRUE`
stretches every line but a paragraph's last to the full width, and
`line_break = "optimal"` chooses a paragraph's breaks together, as TeX
does:

```{r justify, fig.width = 6, fig.height = 1.05, out.width = "90%"}
para <- r"(Least squares has a closed form, which is why it was the
method of choice long before computers made iterative fitting cheap. It
is also optimal among unbiased linear estimators when the errors are
uncorrelated with equal variance: the Gauss--Markov theorem.)"
grid.newpage()
grid.latex(para, input_mode = "document", max_width = 5.6 * 72,
           justify = TRUE, line_break = "optimal",
           x = 0.03, y = 0.95, hjust = 0, vjust = 1,
           gp = gpar(fontsize = 11))
```

## Saving a PDF

A document is one grob, so it goes on a page like any other. Use
`cairo_pdf()` rather than `pdf()`: its text stays text, which can be
selected and searched, where base `pdf()` falls back to drawing glyphs
as outlines. `latex_dims()` gives the size the body needs, to size the
page from:

```{r pdf, eval = FALSE}
d <- latex_dims(body, input_mode = "document", max_width = 5.5 * 72,
                gp = gpar(fontsize = 11))
height <- as.numeric(d$height + d$depth) / 72 + 1   # inches, with margins

cairo_pdf("methods.pdf", width = 6.5, height = height)
grid.latex(body, input_mode = "document", max_width = 5.5 * 72,
           x = unit(0.5, "in"), y = unit(1, "npc") - unit(0.5, "in"),
           hjust = 0, vjust = 1, gp = gpar(fontsize = 11))
dev.off()
```

A body longer than the page does not flow onto the next one: a grob is
one piece. Split a long document at its paragraph breaks and draw each
part on a page of its own.

## What a single grob cannot do

A document's cross-references need the rest of the document, so they
warn and are drawn as LaTeX draws them when it cannot resolve one:
`\ref` and `\pageref` as a bold `??`, `\eqref` as `(??)` and `\cite{key}`
as `[?]`. A `\footnote` has no foot of a page to go to, so its text is set
where it is written. The preamble, `\maketitle` and `\label` are read and
dropped, and a `table` or `figure` is set where it is written, as a float
placed "here" would be. `vignette("getting-started")` lists what is not
supported at all.
