---
title: "Typesetting a document"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Typesetting a document}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "png"
)
library(gridmicrotex)
library(grid)
```

With `input_mode = "document"`, `grid.latex()` sets the body of a LaTeX
document: paragraphs, headings, displayed equations, lists and tables.
Use it for a methods section, an appendix or a long caption, with no
LaTeX installation.

## A document body

This is ordinary LaTeX, as it would appear between `\begin{document}` and
`\end{document}`:

```{r body}
body <- r"(\section{Methods}
We fit a straight line to $n$ observations $(x_i, y_i)$ by least
squares, which chooses the intercept and slope that minimise the sum
of squared residuals
\[ S(\beta_0, \beta_1) = \sum_{i=1}^{n} (y_i - \beta_0 - \beta_1 x_i)^2. \]
Setting both partial derivatives to zero gives the estimates
\begin{align*}
  \hat\beta_1 &= \frac{\sum_i (x_i - \bar x)(y_i - \bar y)}{\sum_i (x_i - \bar x)^2}, \\
  \hat\beta_0 &= \bar y - \hat\beta_1 \bar x.
\end{align*}

The fitted values are $\hat y_i = \hat\beta_0 + \hat\beta_1 x_i$, and
the residuals $e_i = y_i - \hat y_i$ are what is left over.

\subsection{Assumptions}
The errors are taken to be
\begin{itemize}
  \item independent of one another,
  \item of constant variance $\sigma^2$, and
  \item normally distributed --- which matters only for the tests.
\end{itemize}
)"
```

`max_width` is the text width in big points (1/72 inch):

```{r draw-body, fig.width = 6, fig.height = 5.2, out.width = "90%"}
grid.newpage()
grid.latex(body, input_mode = "document", max_width = 5.6 * 72,
           x = 0.03, y = 0.97, hjust = 0, vjust = 1,
           gp = gpar(fontsize = 11))
```

## How the body is read

As in LaTeX:

- **Paragraphs.** A blank line or `\par` starts a new, indented
  paragraph. `\noindent` suppresses the indent.
- **Headings.** `\section`, `\subsection` and `\subsubsection` are
  numbered; starred forms are not.
- **Math.** `$...$` and `\(...\)` are inline. `\[...\]`, `$$...$$`,
  `equation`, `align`, `gather` and `multline` are centred on their own
  line. Equations are not numbered.
- **Lists and tables.** `itemize`, `enumerate` and `tabular` hold text,
  with math between `$...$`.
- **Text.** `--` and `---` are dashes, quotes are curly, and `~` is a
  non-breaking space. `\centering` and `center` centre lines.

Problems produce one warning, listing each with its line and column.

## Justified text

`justify = TRUE` fills every line but the last, and
`line_break = "optimal"` chooses the breaks for the whole paragraph:

```{r justify, fig.width = 6, fig.height = 1.05, out.width = "90%"}
para <- r"(Least squares has a closed form, which is why it was the
method of choice long before computers made iterative fitting cheap. It
is also optimal among unbiased linear estimators when the errors are
uncorrelated with equal variance: the Gauss--Markov theorem.)"
grid.newpage()
grid.latex(para, input_mode = "document", max_width = 5.6 * 72,
           justify = TRUE, line_break = "optimal",
           x = 0.03, y = 0.95, hjust = 0, vjust = 1,
           gp = gpar(fontsize = 11))
```

## Saving a PDF

Use `cairo_pdf()` so the text can be selected. `latex_dims()` gives the
height to size the page:

```{r pdf, eval = FALSE}
d <- latex_dims(body, input_mode = "document", max_width = 5.5 * 72,
                gp = gpar(fontsize = 11))
height <- as.numeric(d$height) / 72 + 1   # inches, with margins

cairo_pdf("methods.pdf", width = 6.5, height = height)
grid.latex(body, input_mode = "document", max_width = 5.5 * 72,
           x = unit(0.5, "in"), y = unit(1, "npc") - unit(0.5, "in"),
           hjust = 0, vjust = 1, gp = gpar(fontsize = 11))
dev.off()
```

A document is drawn as one piece and does not flow onto a second page.
Split a long one at paragraph breaks and draw each part on its own page.

## Limitations

- `\ref` and `\pageref` draw `??`, `\eqref` draws `(??)` and `\cite`
  draws `[?]`, with a warning.
- A footnote's text is set where it is written.
- Tables and figures are placed where they are written.
- The preamble, `\maketitle` and `\label` draw nothing.
