---
title: "Introduction to gridmicrotex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gridmicrotex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "ragg_png"
)

```

## What is gridmicrotex?

**gridmicrotex** renders LaTeX math equations as native R `grid` graphics
objects (grobs). It uses the [MicroTeX](https://github.com/NanoMichael/MicroTeX)
C++ library as its layout engine --- MicroTeX parses LaTeX, builds the TeX
box model, and computes exact glyph coordinates. The package intercepts
this layout data and maps it to native grid primitives (`pathGrob`,
`segmentsGrob`, `rectGrob`, `textGrob`), producing a `gTree` that works
on any R graphics device at any resolution.

**Key features:**


- No external LaTeX installation required --- MicroTeX is fully embedded
- Resolution-independent vector output on all R devices (PNG, PDF, SVG, ...)
- Full math support: fractions, roots, integrals, matrices, Greek letters,
  accents, delimiters, and more
- Two bundled math fonts: Lete Sans Math (sans-serif, default) and
  STIX Two Math (serif); additional fonts via `load_math_font()`
- Color support via `\textcolor{}`
- ggplot2 integration with `geom_latex()` and `element_latex()`
- CJK and multilingual text in `\text{}` blocks
- Markdown rendering with inline math, via `markdown_grob()` and
  `markdown_box_grob()` --- see `vignette("markdown")`

## Basic usage

The core function is `latex_grob()`, which returns a grid grob:

```{r basic, fig.height=0.8, fig.width=2, out.width="50%"}
library(gridmicrotex)
library(grid)

g <- latex_grob("\\frac{-b \\pm \\sqrt{b^2 - 4ac}}{2a}", gp = grid::gpar(fontsize = 24))
grid::grid.newpage()
grid::grid.draw(g)
```

For quick rendering, use `grid.latex()`:

```{r quick, fig.height=0.8, fig.width=3, out.width="40%"}
grid::grid.newpage()
grid.latex("$\\sum_{i=1}^{n} x_i^2$", gp = grid::gpar(fontsize = 28))
```

## Positioning and justification

Control placement with `x`, `y`, `hjust`, and `vjust`:

```{r positioning, fig.height=1.2, fig.width=3.5, out.width="40%"}
grid::grid.newpage()
grid.latex("Famous $E = mc^2$", x = 0.1, y = 0.7, hjust = 0, gp = grid::gpar(fontsize = 24))
grid.latex("F = ma", x = 0.1, y = 0.3, hjust = 0, gp = grid::gpar(fontsize = 24), input_mode = "math")
```

The default `input_mode = "mixed"` reads the string as text and renders
only what sits inside `$…$` or `\(…\)` as math --- which is why
`"Famous"` above needs no `\text{}`. It saves typing, but the conversion
is heuristic. `input_mode = "math"` treats the whole string as math, so
prose must be wrapped in `\text{}`; that is the second example above,
and it suits heavy math or LaTeX macros. Set it for a session with
`latex_options(input_mode = "math")`, as the rest of this vignette does.

### Aligning to the math baseline

In addition to numeric justifications in `[0, 1]`, `hjust` and `vjust`
also accept named values. The most useful one is `vjust = "baseline"`,
which places the formula's math baseline (not the bbox centre) on the
anchor point — so a formula sits next to surrounding text the way it
would in a typeset document.

```{r baseline-align, fig.height=0.8, fig.width=4.5, out.width="60%"}
grid::grid.newpage()
y <- 0.5
grid::grid.segments(unit(0, "npc"), unit(y, "npc"),
                    unit(1, "npc"), unit(y, "npc"),
                    gp = grid::gpar(col = "grey80"))
grid::grid.text("if ", x = 0.10, y = y, just = c(0, 0.5),
                gp = grid::gpar(fontsize = 16))
grid.latex("$x \\geq \\sqrt{2\\pi}$",
           x = 0.22, y = y,
           hjust = "left", vjust = "baseline",
           gp = grid::gpar(fontsize = 16))
grid::grid.text(", then proceed.", x = 0.62, y = y, just = c(0, 0.5),
                gp = grid::gpar(fontsize = 16))
```

`hjust` accepts `"left"`/`"bbleft"`, `"center"`/`"centre"`/`"middle"`/
`"bbcentre"`, and `"right"`/`"bbright"`. `vjust` accepts `"bottom"`,
`"center"`/`"centre"`/`"middle"`, `"top"`, and `"baseline"`.

### Named anchors with `\mark{}`

The `\mark{name}` macro records a named anchor inside the formula, and
`grobMark()` resolves it to a pair of grid units ready to drive an arrow
or callout. Marks work at any nesting level --- even inside a
superscript or fraction --- and inherit the surrounding transform (font
shrink, scaling, rotation), so the anchor lands on the rendered glyph.

```{r mark, fig.height=2.4, fig.width=5, out.width="70%"}
g <- latex_grob(
  r"($a^2 + b\mark{term}^2 \mark{equals}= c^2$)",
  x = 0.5, y = 0.4,
  gp = grid::gpar(fontsize = 28)
)
grid::grid.newpage()
grid::grid.draw(g)

# Callout 1: the "=" sign, pointed at from above.
mk_eq <- grobMark(g, "equals")
grid::grid.segments(mk_eq$x, mk_eq$y + unit(15, "mm"),
                    mk_eq$x, mk_eq$y + unit(3, "mm"),
                    arrow = grid::arrow(length = unit(2, "mm"), type = "closed"),
                    gp = grid::gpar(col = "red"))
grid::grid.text("equals", x = mk_eq$x, y = mk_eq$y + unit(18, "mm"),
                gp = grid::gpar(col = "red", fontsize = 11))

# Callout 2: the b^2 term, pointed at from below — the mark sits at the
# end of the term, including the superscript's smaller scale.
mk_bsq <- grobMark(g, "term")
grid::grid.segments(mk_bsq$x - unit(6, "mm"), mk_bsq$y - unit(15, "mm"),
                    mk_bsq$x - unit(2, "mm"), mk_bsq$y - unit(3, "mm"),
                    arrow = grid::arrow(length = unit(2, "mm"), type = "closed"),
                    gp = grid::gpar(col = "blue"))
grid::grid.text("b² term",
                x = mk_bsq$x - unit(7, "mm"),
                y = mk_bsq$y - unit(18, "mm"),
                just = "right",
                gp = grid::gpar(col = "blue", fontsize = 11))
```

The returned units carry the grob's viewport position and `hjust`/`vjust`,
so they go straight into any grid drawing function with no offset
arithmetic. A mark is a single point, not a span: to centre a callout over
a multi-glyph term, use a pair (`\mark{l}…\mark{r}`) and take the midpoint.

## Colors

You can use `r"()"` raw strings to write LaTeX with regular newlines and quotes without escaping. Set the formula color via `gp`, or use `\textcolor{}` within the LaTeX:

```{r colors, fig.height=0.6, fig.width=2, out.width="40%"}
latex_options(input_mode = "math") # Set math mode globally
grid::grid.newpage()
grid.latex(
  r"(\textcolor{red}{\alpha} + \textcolor{blue}{\beta} = \gamma)",
  gp = grid::gpar(fontsize = 28)
)
```

## Math fonts

The package ships with two math fonts, both loaded automatically:

| Alias | Font | License |
|-------|------|---------|
| `"lete"` (default) | Lete Sans Math | SIL Open Font License |
| `"stix"` | STIX Two Math | SIL Open Font License |

```{r fonts-list}
available_math_fonts()
```

Set one per call with `math_font`, or for the session with
`latex_options(math_font = )`:

```{r fonts, fig.height=1.5, fig.width=2, out.width="50%"}
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(2, 1)))
grid::pushViewport(grid::viewport(layout.pos.row = 1))
grid.latex(r"(\int_0^1 f(x)\,dx)", gp = grid::gpar(fontsize = 24))
grid::upViewport()
grid::pushViewport(grid::viewport(layout.pos.row = 2))
grid.latex(r"(\int_0^1 f(x)\,dx)", gp = grid::gpar(fontsize = 24), math_font = "stix")
grid::upViewport(2)
```

`check_math_fonts()` gives a diagnostic report.

### Advanced: loading custom math fonts

Use `load_math_font()` to add any additional OpenType math font. The
OpenType MATH table is parsed directly in C++ --- no companion metrics
file or external toolchain is required:

```r
load_math_font("path/to/MyFont.otf")
```

This is only for **math** fonts. Text fonts --- the ones used inside
`\text{}` blocks --- need no loading at all: set `gp$fontfamily`, or
name one for a single run with `\gmfontfamily{}{}` (see *Naming a font
for one run* above).

### Render modes

- **`"typeface"`** (default) draws glyphs as native text, so PDF/SVG
  output stays selectable and searchable. Fonts are read straight from
  their OTF files --- no system-wide install needed --- but it requires
  a device with the R \eqn{\geq} 4.3 glyph engine (`ragg`, `svglite`,
  `cairo_pdf`). On others, such as the base `pdf()` device, it falls
  back to path mode with a warning.

- **`"path"`** draws each glyph as a filled vector path. Works on every
  device; the text is not selectable.

```r
# Default typeface mode (selectable text in PDF/SVG)
grid.latex("E = mc^2", gp = grid::gpar(fontsize = 24))

# Explicit path mode (works everywhere, but text is not selectable)
grid.latex("E = mc^2", gp = grid::gpar(fontsize = 24), render_mode = "path")
```

> **Do not use `showtext::showtext_auto()` with typeface mode.** showtext
> intercepts all text rendering and converts it to paths, silently
> defeating typeface mode even on `svglite` and `ragg`. Call
> `showtext::showtext_auto(FALSE)` before drawing formulas.

## Querying dimensions

`latex_dims()` returns the bounding box of an expression, for layout
calculations and checking that labels fit:

```{r dims}
latex_dims("\\frac{a}{b}", gp = grid::gpar(fontsize = 20))
```

## Text rendering and CJK support

Text inside `\text{}` and `\mbox{}` goes through R's own text rendering,
so `gp$fontfamily` controls **all** text content --- Latin, CJK, Cyrillic,
and any other script your device supports:

```{r example-mixed-cjk-math, fig.height=0.8, fig.width=5, out.width="50%"}
grid.newpage()
grid.latex(r"(\text{如果 } x > 0 \text{ 则 } y = x^2)",
           gp = gpar(fontsize = 24, fontfamily = "sans"))
```

Any font available to R works: base families like `"sans"`, `"serif"`,
`"mono"`, or fonts registered via **showtext** / **systemfonts**.

### Font pairing

The bundled math fonts have different styles. For a consistent look,
pair them with a matching `fontfamily`:

| Math font | Style | Suggested `fontfamily` |
|-----------|-------|------------------------|
| Lete Sans Math (`"lete"`, default) | Sans-serif | `"sans"` |
| STIX Two Math (`"stix"`) | Serif | `"serif"` |

```{r font-pairing, fig.height = 0.4, fig.width = 2, out.width="70%"}
grid::grid.newpage()
grid.latex(
  "\\text{Theorem: } \\forall x \\in \\mathbb{R},\\; x^2 \\geq 0",
  math_font = "stix",
  gp = grid::gpar(fontfamily = "serif", fontsize = 12)
)
```

### Naming a font for one run

`gp$fontfamily` applies to the whole grob. To put a *single run* of text
in a different font, use `\gmfontfamily{family}{content}` --- a
gridmicrotex extension, not standard LaTeX:

```{r gmfontfamily, fig.height = 0.4, fig.width = 3, out.width="70%"}
grid::grid.newpage()
grid.latex(
  r"(\gmfontfamily{serif}{serif} \quad \gmfontfamily{mono}{mono} \quad \text{default})",
  gp = grid::gpar(fontsize = 16)
)
```

`family` is anything you could pass to `gp$fontfamily`: a generic family
(`"sans"`, `"serif"`, `"mono"`) or a specific name such as `"Georgia"`,
which `\textsf{…}` / `\texttt{…}` cannot express. Unresolvable names fall
back silently. The content is typeset as text, so this styles prose, not
math --- math glyphs follow `math_font`. It composes with emphasis in
either nesting order, but a nested `\gmfontfamily` *replaces* the family
it sits inside.

To go back, use `\textrm{…}`, which returns its content to
`gp$fontfamily` --- including inside a `\textsf{…}`, `\texttt{…}` or
`\gmfontfamily{…}{…}` group, and without disturbing bold or italic:

```{r textrm, fig.height = 0.4, fig.width = 3.5, out.width="80%"}
grid::grid.newpage()
grid.latex(
  r"(\textsf{sans \textrm{back to body} sans})",
  gp = grid::gpar(fontsize = 16, fontfamily = "serif")
)
```

The name is deliberately not `\fontfamily`: LaTeX's takes one argument,
does nothing until `\selectfont`, and wants an NFSS code (`ptm`) rather
than a font name. Since gridmicrotex accepts pasted LaTeX, claiming that
name would silently misparse real input. This is also what markdown's
`font-family` CSS compiles to --- see `vignette("markdown")`.

## Supported LaTeX

gridmicrotex uses the MicroTeX engine, which is a **math formula renderer**,
not a full document typesetter. It covers the vast majority of math notation
you would use in plots and figures, but does not attempt to replace a full
LaTeX installation.

### Complicated examples

Coloured `array` tables, with `\multicolumn`, `\rowcolor`, `\cellcolor`
and nested matrices:

```{r table-multicolumn, fig.height = 3, fig.width=8, out.width="60%"}
grid::grid.newpage()

grid.latex(
  "
  \\newcolumntype{s}{>{\\color{#1234B6}}c}
\\begin{array}{|c|c|c|s|}
  \\hline
  \\rowcolor{Tan}\\multicolumn{4}{|c|}{\\textcolor{white}{\\bold{\\text{Table Head}}}}\\\\
  \\hline
  \\text{Matrix}&\\multicolumn{2}{|c|}{\\text{Multicolumns}}&\\text{Font size commands}\\\\
  \\hline
  \\begin{pmatrix}
      \\alpha_{11}&\\cdots&\\alpha_{1n}\\\\
      \\hdotsfor{3}\\\\
      \\alpha_{n1}&\\cdots&\\alpha_{nn}
  \\end{pmatrix}
  &\\large \\text{Left}&\\cellcolor{#00bde5}\\small \\textcolor{white}{\\text{\\bold{Right}}}
  &\\small \\text{small Small}\\\\
  \\hline
  \\multicolumn{4}{|c|}{\\text{Table Foot}}\\\\
  \\hline
\\end{array}
  ",
  gp = grid::gpar(fontsize = 22)
)
```

```{r complicated-equation, fig.height = 9, fig.width=9, out.width="60%"}
grid::grid.newpage()
grid.latex(
  "\\definecolor{gris}{gray}{0.9}
\\definecolor{noir}{rgb}{0,0,0}
\\definecolor{bleu}{rgb}{0,0,1}
\\fatalIfCmdConflict{false}
\\newcommand{\\pa}{\\left|}
\\begin{array}{c}
  \\LaTeX\\\\
  \\begin{split}
      |I_2| &= \\pa\\int_0^T\\psi(t)\\left\\{ u(a,t)-\\int_{\\gamma(t)}^a \\frac{d\\theta}{k} (\\theta,t) \\int_a^\\theta c(\\xi)
          u_t (\\xi,t)\\,d\\xi\\right\\}dt\\right|\\\\
      &\\le C_6 \\Bigg|\\pa f \\int_\\Omega \\pa\\widetilde{S}^{-1,0}_{a,-}
          W_2(\\Omega, \\Gamma_1)\\right|\\ \\right|\\left| |u|\\overset{\\circ}{\\to} W_2^{\\widetilde{A}}(\\Omega\\Gamma_r,T)\\right|\\Bigg|\\\\
      &\\\\
      &\\begin{pmatrix}
          \\alpha&\\beta&\\gamma&\\delta\\\\
          \\aleph&\\beth&\\gimel&\\daleth\\\\
          \\mathfrak{A}&\\mathfrak{B}&\\mathfrak{C}&\\mathfrak{D}\\\\
          \\boldsymbol{\\mathfrak{a}}&\\boldsymbol{\\mathfrak{b}}&\\boldsymbol{\\mathfrak{c}}&\\boldsymbol{\\mathfrak{d}}
      \\end{pmatrix}
      \\quad{(a+b)}^{\\frac{n}{2}}=\\sqrt{\\sum_{k=0}^n\\tbinom{n}{k}a^kb^{n-k}}\\quad
          \\Biggl(\\biggl(\\Bigl(\\bigl(()\\bigr)\\Bigr)\\biggr)\\Biggr)\\\\
      &\\forall\\varepsilon\\in\\mathbb{R}_+^*\\ \\exists\\eta>0\\ |x-x_0|\\leq\\eta\\Longrightarrow|f(x)-f(x_0)|\\leq\\varepsilon\\\\
      &\\det
      \\begin{bmatrix}
          a_{11}&a_{12}&\\cdots&a_{1n}\\\\
          a_{21}&\\ddots&&\\vdots\\\\
          \\vdots&&\\ddots&\\vdots\\\\
          a_{n1}&\\cdots&\\cdots&a_{nn}
      \\end{bmatrix}
      \\overset{\\mathrm{def}}{=}\\sum_{\\sigma\\in\\mathfrak{S}_n}\\varepsilon(\\sigma)\\prod_{k=1}^n a_{k\\sigma(k)}\\\\
      &\\Delta f(x,y)=\\frac{\\partial^2f}{\\partial x^2}+\\frac{\\partial^2f}{\\partial y^2}\\qquad\\qquad \\fcolorbox{noir}{gris}
          {n!\\underset{n\\rightarrow+\\infty}{\\sim} {\\left(\\frac{n}{e}\\right)}^n\\sqrt{2\\pi n}}\\\\
      &\\sideset{_\\alpha^\\beta}{_\\gamma^\\delta}{
      \\begin{pmatrix}
          a&b\\\\
          c&d
      \\end{pmatrix}}
      \\xrightarrow[T]{n\\pm i-j}\\sideset{^t}{}A\\xleftarrow{\\overrightarrow{u}\\wedge\\overrightarrow{v}}
          \\underleftrightarrow{\\iint_{\\mathds{R}^2}e^{-\\left(x^2+y^2\\right)}\\,\\mathrm{d}x\\mathrm{d}y}
  \\end{split}\\\\
  \\rotatebox{30}{\\sum_{n=1}^{+\\infty}}\\quad\\mbox{Mirror rorriM}\\reflectbox{\\mbox{Mirror rorriM}}
\\end{array}",
  gp = grid::gpar(fontsize = 22),
  render_mode = "path"
)
```



### Lists

The `itemize` and `enumerate` environments lay their items out as a
left-aligned column, one item per row --- `itemize` prefixes each item
with a bullet, `enumerate` numbers them:

```{r lists, fig.height=1.4, fig.width=3, out.width="45%"}
grid::grid.newpage()
grid.latex(paste0(
  "\\begin{enumerate}",
  "  \\item e^{i\\pi} + 1 = 0",
  "  \\item \\begin{itemize}",
  "           \\item \\alpha \\item \\beta",
  "         \\end{itemize}",
  "\\end{enumerate}"
), gp = grid::gpar(fontsize = 20))
```

An optional `[…]` argument customises the marker. For `itemize` it is
the literal marker (`\begin{itemize}[\star]`); for `enumerate` it is a
counter template containing one of `\arabic*`, `\alph*`, `\Alph*`,
`\roman*`, or `\Roman*` (e.g. `\begin{enumerate}[\Roman*.]`). Lists may
nest, and an item may contain any math --- including a `\begin{array}`
table.

Because MicroTeX is a math engine, each item is typeset as a
**math-mode, single-line** expression: there is no paragraph flow or
line wrapping, and prose inside an item must be wrapped in `\text{}`
(e.g. `\item \text{First point}`). The `description` environment is not
supported.

### Pasting LaTeX from other sources

When the input is generated by tools that emit complete LaTeX --- for
example R packages that produce ready-to-compile `tabular` snippets, or
LaTeX fragments copy-pasted from a `.tex` file --- the formula often
arrives wrapped in document-level constructs that MicroTeX does not
implement. Rather than refusing such input, gridmicrotex rewrites or
removes a small set of well-known wrappers before parsing.

**Removed silently (no visual effect):**

| Construct | Why |
|---|---|
| `%`-to-end-of-line comments (`\%` is preserved) | comments are non-visual in LaTeX too |
| `\documentclass[…]{…}`, `\usepackage[…]{…}` | preamble metadata |
| `\begin{document}` / `\end{document}` | document boundary, structural only |
| `\maketitle`, `\title{…}`, `\author{…}` | title-page metadata, no body output |
| `\label{…}` | cross-reference target, never rendered in LaTeX either |
| `\begin{table}[…]` / `\end{table}`, `\begin{figure}[…]` / `\end{figure}` (and starred variants) | float wrappers; the contents stay |
| `\centering`, `\raggedright`, `\raggedleft`, `\flushleft`, `\flushright` | alignment scope declarations |
| `\noindent`, `\relax` | content-free declarations |

**Rewritten to a MicroTeX equivalent:**

| Construct | Becomes |
|---|---|
| `\emph{X}` | `\textit{X}` |
| `\textnormal{X}` | `\text{X}` |
| `\par`, `\newline` | `\\` (line break) |
| `\toprule`, `\bottomrule` | `\thickhline` (rendered ~2× thickness) |
| `\midrule` | `\hline` |
| `\cmidrule[trim]?(parenarg)?{a-b}` | `\cline{a-b}` --- partial-column rule |
| `\caption[short]{X}` | `\text{X}\\` inserted inline at source position |
| `\smallskip`, `\medskip`, `\bigskip` | `\vspace{0.25em}` / `\vspace{0.5em}` / `\vspace{1em}` --- em-relative so they scale with `gp$fontsize` |
| `\hfill`, `\vfill` | `\quad` / `\vspace{1em}` --- static proxies for rubber lengths |

Three of those rewrites are approximations. `\caption` renders where it
appears in the source, not where LaTeX's float machinery would move it
--- so caption-above or caption-below follows whatever your tool emits.
The skips are em-relative rather than LaTeX's absolute 3/6/12 pt, so
they stay visible at any `gp$fontsize`; use `\vspace{Xpt}` for an exact
amount. And `\hfill` / `\vfill` are *rubber* lengths with nothing to
fill in a fixed-size grob, so they become a static 1em gap --- right
position, no elasticity.

**Not honored (rendered as literal text, which is intentional --- it
makes unsupported markup easy to spot):**

- Declarative font scopes: `\bfseries`, `\itshape`, `\ttfamily`,
  `\sffamily`, `\rmfamily`. These affect text within their group in
  LaTeX, which requires scope tracking we do not implement. Use the
  argument-bearing forms instead: `\textbf{…}`, `\textit{…}`,
  `\texttt{…}`, `\textsf{…}`, `\textrm{…}`, all of which nest with each
  other. To choose the text font itself, set `gp$fontfamily`, or name one
  for a single run with `\gmfontfamily{…}{…}` --- see *Naming a font for
  one run* above.
- Hyperlinks and references: `\url{…}`, `\href{…}{…}`, `\ref{…}`,
  `\cite{…}`. There is no link concept inside a grob.
- Footnotes: `\footnote{…}` (positioning machinery is page-bound).
- Small caps: `\textsc{…}` (MicroTeX has no small-caps glyphs).

### What is not supported

These need a real document compiler and are outside a formula renderer's
scope: document structure (`\section`, page layout, `\tableofcontents`);
paragraph text (line breaking, hyphenation, justification); TikZ/PGF;
`\includegraphics`; cross-references and bibliographies; theorem
environments; the `description` list environment (`itemize` and
`enumerate` *are* supported); and `\tag` / equation numbering.
`\usepackage{…}` is accepted but loads nothing --- every supported
command is built into MicroTeX.

For axis labels, annotations, legends and in-plot formulas, the
supported set is more than sufficient.

## Project-wide defaults

`latex_options()` sets `math_font`, `render_mode` and `input_mode` for
calls that don't supply them; explicit arguments always win. Size stays
at the grob level via `gp$fontsize` / `gp$lineheight`.

```{r options, eval=FALSE}
latex_options(math_font = "stix", render_mode = "typeface")
latex_options()        # query
reset_latex_options()  # back to built-in defaults
```

## User-defined macros

`define_macro()` registers zero-argument shorthands that are expanded
by text substitution before the expression reaches MicroTeX. Handy for
recurring notation:

```{r macros, fig.height=0.7, fig.width=3, out.width="50%"}
define_macro("RR", "\\mathbb{R}")
define_macro("eps", "\\varepsilon")

grid::grid.newpage()
grid.latex("\\forall \\eps > 0, \\eps \\in \\RR", gp = grid::gpar(fontsize = 24))

clear_macros()
```

Macro names must be ASCII letters, and expansion iterates to a fixed
point so macros can reference each other. `list_macros()` shows what is
registered; `clear_macros()` drops everything.

For parameterised macros (0--9 arguments), MicroTeX also accepts
plain-TeX `\def`. These live only for the expression they appear in, so
use them for an abbreviation local to one label and `define_macro()` for
one that should persist across plots:

```{r def-inline, fig.height=0.7, fig.width=4, out.width="60%"}
grid::grid.newpage()
grid.latex(
  r"(\def\norm#1{\left\lVert #1 \right\rVert}
      \norm{\vec{v}} = \sqrt{\langle \vec{v}, \vec{v} \rangle})",
  gp = grid::gpar(fontsize = 24)
)
```

## Caching and introspection

Parsed layouts are memoised by `(tex, fontsize, math_font, render_mode,
…)`, so a repeated axis label is only laid out once:

```{r cache, eval=FALSE}
latex_cache_info()       # size / max_size / hits / misses
latex_cache_limit(1024)  # LRU capacity; 0 disables caching
latex_cache_clear()      # wipe (e.g. after re-loading fonts)
```

`latex_tree()` returns the raw draw records plus bbox metadata, and
`debug = TRUE` overlays the bounding box, baseline and record origins on
the rendered formula --- both useful when checking alignment:

```{r debug, fig.height=1, fig.width=3, out.width="60%"}
grid::grid.newpage()
grid.latex("x^{2} + y_{i}", gp = grid::gpar(fontsize = 30), debug = TRUE)
```

## Graphics backend

Use `ragg::agg_png()`, `svglite::svglite()` or `grDevices::cairo_pdf()`.
The default devices on Windows and macOS may not find the bundled math
fonts, warning `font family not found in Windows font database`. The
README covers the setup, and a comparison with `tikzDevice`, `xdvir`,
`latex2exp` and `plotmath`.

