---
title: "Introduction to gridmicrotex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gridmicrotex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "png"
)
```

## What is gridmicrotex?

**gridmicrotex** renders LaTeX math as native R `grid` graphics objects,
embedding the [MicroTeX](https://github.com/NanoMichael/MicroTeX) C++
engine. The pipeline is:

1. MicroTeX parses the LaTeX into a TeX box model
2. A custom `Graphics2D` recorder captures every draw operation --- glyph
   paths, lines, rectangles --- with exact coordinates
3. That layout crosses the C++/R boundary as a data frame
4. R converts each record into a grid primitive (`pathGrob`,
   `segmentsGrob`, `rectGrob`, `textGrob`)
5. The result is a `gTree` that draws on any device at any resolution

No LaTeX installation is required, and because nothing is rasterised the
output is resolution independent on every R device.

**Base graphics** reuses steps 1--3 and diverges at step 4: with
`latex_options(device_math = TRUE)` the same layout is emitted straight
through the graphics device's own drawing callbacks instead of being
turned into grobs, which is how `plot()` titles and axis labels get real
LaTeX without any change to your code. Because that interception sits at
the device, it is not limited to base graphics: text drawn by grid,
ggplot2 and lattice passes through it too. See
[Base graphics](#base-graphics) below.

- Full math: fractions, roots, integrals, matrices, Greek, accents,
  delimiters, and colour via `\textcolor{}`
- Two bundled math fonts, plus any OpenType math font through
  `load_math_font()`
- CJK, RTL and multilingual text inside `\text{}`
- Base graphics --- `latex_options(device_math = TRUE)`
- ggplot2 integration --- `geom_latex()` and `element_latex()`, see
  `vignette("ggplot2-integration")`
- Markdown with inline math --- see `vignette("markdown")`

## Quick start

`latex_grob()` returns a grob; `grid.latex()` builds and draws one.

```{r basic, fig.height=1, fig.width=2.4, out.width="45%"}
library(gridmicrotex)
library(grid)

grid.newpage()
grid.latex(r"($\frac{\textcolor{red}{-b} \pm \sqrt{b^2 - 4ac}}{2a}$)")
```

**Write LaTeX in a raw string.** `r"(...)"` passes backslashes through
untouched, so what you paste is what renders. Every example here uses
one. In an ordinary `"..."` string each `\` has to be doubled --- and
LaTeX's own row separator `\\` becomes a bewildering `\\\\`.

### Mixing text and math

The default `input_mode = "mixed"` reads the string as prose and
typesets only what sits inside `$…$` or `\(…\)` as math, which is why
`Famous:` below needs no markup. It saves typing, but the split is
heuristic. `input_mode = "math"` treats the whole string as math, so
prose must be wrapped in `\text{}`; that suits heavy math and pasted
LaTeX. These two render identically:

```{r modes, fig.height=1.2, fig.width=3.2, out.width="45%"}
grid.newpage()
grid.latex(r"(Famous: $E = mc^2$)",
           x = 0.05, y = 0.7, hjust = 0, )
grid.latex(r"(\text{Famous: } E = mc^2)", input_mode = "math",
           x = 0.05, y = 0.3, hjust = 0, )
```

Set the mode for a whole session with `latex_options(input_mode = )`.

## What it can render

Two figures before the details, because the range is easier to see than
to describe. A coloured `array` with `\multicolumn`, `\rowcolor`,
`\cellcolor`, a custom column type and a nested matrix:

```{r showcase-table, fig.height = 3, fig.width = 8, out.width="70%"}
grid.newpage()
grid.latex(r"(
\newcolumntype{s}{>{\color{#1234B6}}c}
\begin{array}{|c|c|c|s|}
  \hline
  \rowcolor{Tan}\multicolumn{4}{|c|}{\textcolor{white}{\bold{\text{Table Head}}}}\\
  \hline
  \text{Matrix}&\multicolumn{2}{|c|}{\text{Multicolumns}}&\text{Font size commands}\\
  \hline
  \begin{pmatrix}
      \alpha_{11}&\cdots&\alpha_{1n}\\
      \hdotsfor{3}\\
      \alpha_{n1}&\cdots&\alpha_{nn}
  \end{pmatrix}
  &\large \text{Left}&\cellcolor{#00bde5}\small \textcolor{white}{\text{\bold{Right}}}
  &\small \text{small Small}\\
  \hline
  \multicolumn{4}{|c|}{\text{Table Foot}}\\
  \hline
\end{array}
)", )
```

And a page of assorted notation --- `split` alignment, fraktur, stacked
delimiters, `\sideset`, extensible arrows, `\rotatebox`, `\reflectbox`
and a boxed result:

```{r showcase-equation, fig.height = 9, fig.width = 9, out.width="70%"}
grid.newpage()
grid.latex(r"(
\definecolor{gris}{gray}{0.9}
\definecolor{noir}{rgb}{0,0,0}
\fatalIfCmdConflict{false}
\newcommand{\pa}{\left|}
\begin{array}{c}
  \LaTeX\\
  \begin{split}
      |I_2| &= \pa\int_0^T\psi(t)\left\{ u(a,t)-\int_{\gamma(t)}^a \frac{d\theta}{k} (\theta,t) \int_a^\theta c(\xi)
          u_t (\xi,t)\,d\xi\right\}dt\right|\\
      &\le C_6 \Bigg|\pa f \int_\Omega \pa\widetilde{S}^{-1,0}_{a,-}
          W_2(\Omega, \Gamma_1)\right|\ \right|\left| |u|\overset{\circ}{\to} W_2^{\widetilde{A}}(\Omega\Gamma_r,T)\right|\Bigg|\\
      &\\
      &\begin{pmatrix}
          \alpha&\beta&\gamma&\delta\\
          \aleph&\beth&\gimel&\daleth\\
          \mathfrak{A}&\mathfrak{B}&\mathfrak{C}&\mathfrak{D}\\
          \boldsymbol{\mathfrak{a}}&\boldsymbol{\mathfrak{b}}&\boldsymbol{\mathfrak{c}}&\boldsymbol{\mathfrak{d}}
      \end{pmatrix}
      \quad{(a+b)}^{\frac{n}{2}}=\sqrt{\sum_{k=0}^n\tbinom{n}{k}a^kb^{n-k}}\quad
          \Biggl(\biggl(\Bigl(\bigl(()\bigr)\Bigr)\biggr)\Biggr)\\
      &\forall\varepsilon\in\mathbb{R}_+^*\ \exists\eta>0\ |x-x_0|\leq\eta\Longrightarrow|f(x)-f(x_0)|\leq\varepsilon\\
      &\det
      \begin{bmatrix}
          a_{11}&a_{12}&\cdots&a_{1n}\\
          a_{21}&\ddots&&\vdots\\
          \vdots&&\ddots&\vdots\\
          a_{n1}&\cdots&\cdots&a_{nn}
      \end{bmatrix}
      \overset{\mathrm{def}}{=}\sum_{\sigma\in\mathfrak{S}_n}\varepsilon(\sigma)\prod_{k=1}^n a_{k\sigma(k)}\\
      &\Delta f(x,y)=\frac{\partial^2f}{\partial x^2}+\frac{\partial^2f}{\partial y^2}\qquad\qquad \fcolorbox{noir}{gris}
          {n!\underset{n\rightarrow+\infty}{\sim} {\left(\frac{n}{e}\right)}^n\sqrt{2\pi n}}\\
      &\sideset{_\alpha^\beta}{_\gamma^\delta}{
      \begin{pmatrix}
          a&b\\
          c&d
      \end{pmatrix}}
      \xrightarrow[T]{n\pm i-j}\sideset{^t}{}A\xleftarrow{\overrightarrow{u}\wedge\overrightarrow{v}}
          \underleftrightarrow{\iint_{\mathds{R}^2}e^{-\left(x^2+y^2\right)}\,\mathrm{d}x\mathrm{d}y}
  \end{split}\\
  \rotatebox{30}{\sum_{n=1}^{+\infty}}\quad\mbox{Mirror rorriM}\reflectbox{\mbox{Mirror rorriM}}
\end{array}
)", render_mode = "path")
```

Both are ordinary LaTeX, pasted unchanged. The rest of this vignette is
about placing such things, choosing fonts, and knowing where the
supported set ends.


### Aligning to the math baseline

`hjust` and `vjust` also take names. The useful one is
`vjust = "baseline"`, which puts the formula's *math baseline* --- not
the bounding-box centre --- on the anchor, so a formula sits beside
running text the way it would in a typeset document.

```{r baseline-align, fig.height=1.1, fig.width=6, out.width="70%"}
grid.newpage()
y <- 0.5
grid.segments(unit(0, "npc"), unit(y, "npc"),
              unit(1, "npc"), unit(y, "npc"), gp = gpar(col = "grey80"))
grid.text("if ", x = 0.10, y = y, just = c(0, 0.5), gp = gpar(fontsize = 20))
grid.latex(r"($x \geq \sqrt{2\pi}$)",
           x = 0.22, y = y, hjust = "left", vjust = "baseline",
           gp = gpar(fontsize = 20))
grid.text(", then proceed.", x = 0.62, y = y, just = c(0, 0.5),
          gp = gpar(fontsize = 20))
```

`hjust` accepts `"left"`/`"bbleft"`, `"center"`/`"centre"`/`"middle"`/
`"bbcentre"`, and `"right"`/`"bbright"`. `vjust` accepts `"bottom"`,
`"center"`/`"centre"`/`"middle"`, `"top"`, and `"baseline"`.

### Named anchors with `\mark{}`

`\mark{name}` records a named anchor inside the formula, and `grobMark()`
resolves it to a pair of grid units ready to drive an arrow or a callout.
Marks work at any nesting level --- even inside a superscript or a
fraction --- and inherit the surrounding transform (font shrink, scaling,
rotation), so the anchor lands on the rendered glyph.

```{r mark, fig.height=2.4, fig.width=5, out.width="70%"}
g <- latex_grob(r"($a^2 + b\mark{term}^2 \mark{equals}= c^2$)",
                x = 0.5, y = 0.4)
grid.newpage()
grid.draw(g)

# The "=" sign, pointed at from above.
mk_eq <- grobMark(g, "equals")
grid.segments(mk_eq$x, mk_eq$y + unit(15, "mm"),
              mk_eq$x, mk_eq$y + unit(3, "mm"),
              arrow = arrow(length = unit(2, "mm"), type = "closed"),
              gp = gpar(col = "red"))
grid.text("equals", x = mk_eq$x, y = mk_eq$y + unit(18, "mm"),
          gp = gpar(col = "red"))

# The b^2 term, from below -- the mark sits at the end of the term,
# including the superscript's smaller scale.
mk_bsq <- grobMark(g, "term")
grid.segments(mk_bsq$x - unit(6, "mm"), mk_bsq$y - unit(15, "mm"),
              mk_bsq$x - unit(2, "mm"), mk_bsq$y - unit(3, "mm"),
              arrow = arrow(length = unit(2, "mm"), type = "closed"),
              gp = gpar(col = "blue"))
grid.text("b² term", x = mk_bsq$x - unit(7, "mm"),
          y = mk_bsq$y - unit(18, "mm"), just = "right",
          gp = gpar(col = "blue"))
```

The returned units carry the grob's viewport position and `hjust`/`vjust`,
so they go straight into any grid drawing function with no offset
arithmetic. A mark is a single point, not a span: to centre a callout over
a multi-glyph term, use a pair (`\mark{l}…\mark{r}`) and take the midpoint.

## Display and inline style

TeX sets the same expression two ways. In *display* style a `\sum` is
drawn at full size with its limits above and below; in *text* (inline)
style it shrinks and the limits become scripts beside it. This is a
choice about spacing and layout, not about size --- size is always
`gp$fontsize`.

By default the delimiters decide, exactly as in a LaTeX document: `$…$`
gives text style and `$$…$$` gives display style. A label written without
delimiters --- a plot title, an axis title, an `element_latex()` --- has
nothing for the parser to go on, so it gets text style. `tex_style`
overrides that for the whole expression, and takes all four of TeX's
styles:

```{r style-options, fig.height=1.9, fig.width=6.5, out.width="100%"}
sum_expr <- r"(\sum_{i=1}^{n} \frac{x_i}{n})"
styles <- c("display", "text", "script", "scriptscript")
labels <- c('"display"  ($$...$$)', '"text"  ($...$)',
            '"script"', '"scriptscript"')

grid.newpage()
for (i in seq_along(styles)) {
  pushViewport(viewport(x = (i - 0.5) / 4, width = 1 / 4))
  grid.text(labels[i], y = 0.88, gp = gpar(cex = 0.75, fontface = "bold"))
  grid.latex(sum_expr, y = 0.42, input_mode = "math",
             tex_style = styles[i], gp = gpar(fontsize = 20))
  grid.rect(gp = gpar(col = "grey85", fill = NA))
  popViewport()
}
```

Every panel is `gp = gpar(fontsize = 20)`. The display version is taller
and narrower because the limits moved above and below the operator
instead of sitting beside it --- worth knowing when you are budgeting
space for a label, and `latex_dims()` reports the difference if you need
the numbers. `"script"` and `"scriptscript"` are the sizes TeX uses for
first- and second-level scripts; they are occasionally handy for a
compact sub-label, and rarely needed otherwise.

To change style for part of an expression rather than all of it, use the
inline commands `\displaystyle`, `\textstyle`, `\scriptstyle` and
`\scriptscriptstyle` instead. `latex_options(tex_style = )` sets a
session default.

## Wrapping and justification

`max_width`, in big points, wraps a label over as many lines as it needs.

```{r wrap, fig.height = 2, fig.width = 4, out.width = "65%"}
prose <- paste(rep(
  r"(The quick brown fox jumps over the lazy dog, and $x^2$ too.)", 3),
  collapse = " ")

grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 1)))
pushViewport(viewport(layout.pos.row = 1))
grid.text("ragged (default)", x = 0.02, y = 0.98, hjust = 0, vjust = 1,
          gp = gpar(col = "grey40"))
grid.latex(prose, x = 0.02, y = 0.78, hjust = 0, vjust = 1,
           max_width = 3.6 * 72, gp = gpar(fontsize = 11))
popViewport()
pushViewport(viewport(layout.pos.row = 2))
grid.text("justified + optimal", x = 0.02, y = 0.98, hjust = 0, vjust = 1,
          gp = gpar(col = "grey40"))
grid.latex(prose, x = 0.02, y = 0.78, hjust = 0, vjust = 1,
           max_width = 3.6 * 72, justify = TRUE, line_break = "optimal",
           gp = gpar(fontsize = 11))
popViewport(2)
```

Both refinements are off by default. `justify = TRUE` fills every line but
the last; `line_break = "optimal"` chooses the breaks by total fit rather
than one line at a time. Justifying a narrow column opens wide word gaps,
since nothing may break inside a word --- mark the words that may with
`\-`, as in `in\-ter\-na\-tion\-al`.

## Including a figure

`\includegraphics` draws an image inside the formula. PNG, JPEG and SVG
are supported, sized with `width`, `height` or `scale` in any LaTeX
length; give one of width/height and the aspect ratio is kept, or both
plus `keepaspectratio` to fit inside them. As in LaTeX the extension may
be omitted (`{plots/fig}` finds `plots/fig.svg` or `plots/fig.png`), and
`\graphicspath{{figs/}}` adds a directory to search. `angle` rotates the
figure, like `\rotatebox`; `trim` and `clip` are recognised but not
applied, and say so.

```{r img-basic, fig.height = 1.6, fig.width = 4, out.width = "60%"}
fig <- tempfile(fileext = ".svg")
svglite::svglite(fig, width = 2, height = 1.2)
grid.newpage()
grid.circle(r = 0.35, gp = gpar(fill = "steelblue", col = NA))
dev.off()

grid.newpage()
grid.latex(sprintf(r"(\text{before }\includegraphics[width=1in]{%s}\text{ after})", fig),
           input_mode = "math", gp = gpar(fontsize = 20))
```

Inline, the image sits **on the baseline**, as in LaTeX --- `\raisebox`
moves it. A paragraph holding only an image is a **block**, flush left
unless `halign` or a `text-align` rule on `img` says otherwise.

```{r img-raise, fig.height = 1.5, fig.width = 4.5, out.width = "70%"}
icon <- sprintf(r"(\includegraphics[width=14bp]{%s})", fig)
grid.newpage()
grid.latex(sprintf(r"(\text{baseline }%s\text{, raised }\raisebox{4bp}{%s})",
                   icon, icon),
           input_mode = "math", gp = gpar(fontsize = 20))
```

### A caption

`\caption` renders where it is written. A caption after `\includegraphics`
therefore sits below the figure, and one written before a `tabular` sits
above the table --- which is what makes raw `xtable` and `kable` output
come out the right way round without any float logic.

`\centering` is dropped, because a grob has no page to centre against. A
one-column `array` does the job instead: it centres the figure and its
caption on each other.

```{r img-caption, fig.height = 1.4, fig.width = 3, out.width = "45%"}
logo <- system.file("img", "Rlogo.png", package = "png")
grid.newpage()
grid.latex(sprintf(r"(\begin{array}{c}
  \includegraphics[width=0.6in]{%s}\\
  \caption{Figure 1: the R logo}
\end{array})", logo), gp = gpar(fontsize = 11))
```

Remove the `array` and the same two lines come out flush left. LaTeX
numbers captions from a counter; here the number is yours to write.

**Prefer SVG.** An SVG is drawn as real grid primitives, so it stays sharp
at any output resolution — there is no dpi to choose. A PNG or JPEG is a
fixed grid of pixels, so if you display one much larger than it was saved
you get a blurry result; the package warns when the effective resolution
falls below 150 dpi and tells you what pixel size would reach 300. The
rule is `pixels = display_inches * output_dpi`.

PDF and EPS are **not** supported. Save the figure as SVG instead: it needs
no external tool and, unlike a rasterised PDF, its text stays text.


## Fonts

Two layers are in play. **MicroTeX** chooses the math glyphs and their
metrics; **grid** draws everything inside `\text{}` using `gp$fontfamily`,
so prose follows R's ordinary font handling --- Latin, CJK, Cyrillic and
anything else the device supports.

Two math fonts ship with the package and load automatically:

| Alias | Font | Style | Pairs with |
|-------|------|-------|------------|
| `"lete"` (default) | Lete Sans Math | Sans-serif | `fontfamily = "sans"` |
| `"stix"` | STIX Two Math | Serif | `fontfamily = "serif"` |

```{r fonts-list}
available_math_fonts()
```

Set one per call with `math_font`, or for the session with
`latex_options(math_font = )`. Both rows below are the same formula ---
only the math font and its paired text family differ:

```{r fonts, fig.height=1.5, fig.width=3.4, out.width="60%"}
formula <- r"(Theorem: $\int_0^1 f(x)\,dx \geq 0$)"

grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 1)))
pushViewport(viewport(layout.pos.row = 1))
grid.latex(formula, gp = gpar(fontfamily = "sans"))
upViewport()
pushViewport(viewport(layout.pos.row = 2))
grid.latex(formula, math_font = "stix",
           gp = gpar(fontfamily = "serif"))
upViewport(2)
```

`check_math_fonts()` gives a diagnostic report. Any font available to R
works for the text half --- base families like `"sans"`, `"serif"` and
`"mono"`, or anything registered through **systemfonts**:

```{r cjk, fig.height=1, fig.width=5, out.width="55%", dev = "ragg_png", dev.args = list(), purl = FALSE}
grid.newpage()
grid.latex(r"(如果 $x > 0$ 则 $y = x^2$)",
           gp = gpar(fontfamily = "sans"))
```

Right-to-left scripts work the same way, mixed with Latin included. Two
things do not: wrapping a right-to-left paragraph across lines needs the
optional [FriBidi](https://github.com/fribidi/fribidi) library, and a
math atom between two right-to-left runs --- `\text{…}$x^2$\text{…}` ---
falls back to left-to-right order.

### Naming a font for one run

`gp$fontfamily` applies to the whole grob. For a *single run*, use
`\gmfontfamily{family}{content}` --- a gridmicrotex extension, not
standard LaTeX. `\textrm{…}` goes the other way, returning content to
`gp$fontfamily` even inside a `\textsf{…}`, `\texttt{…}` or
`\gmfontfamily{…}{…}` group, and without disturbing bold or italic:

```{r gmfontfamily, fig.height=0.8, fig.width=4.5, out.width="85%", dev = "ragg_png", dev.args = list()}
grid.newpage()
grid.latex(
  r"(\textsf{sans \textrm{body} sans} \quad \gmfontfamily{mono}{mono})",
  gp = gpar(fontfamily = "serif")
)
```

`family` is anything `gp$fontfamily` accepts: a generic (`"sans"`,
`"serif"`, `"mono"`) or a specific name such as `"Georgia"`, which
`\textsf{…}` / `\texttt{…}` cannot express. Unresolvable names fall back
silently. The content is typeset as text, so this styles prose, not math
--- math glyphs follow `math_font`. It composes with emphasis in either
nesting order, but a nested `\gmfontfamily` *replaces* the family it sits
inside.

The name is deliberately not `\fontfamily`: LaTeX's takes one argument,
does nothing until `\selectfont`, and wants an NFSS code (`ptm`) rather
than a font name. Since gridmicrotex accepts pasted LaTeX, claiming that
name would silently misparse real input. This is also what markdown's
`font-family` CSS compiles to --- see `vignette("markdown")`.

### Loading a custom math font

`load_math_font()` adds any OpenType math font. The OpenType MATH table
is parsed directly in C++, so no companion metrics file and no external
toolchain are needed:

```r
load_math_font("path/to/MyFont.otf")
```

This is only for **math** fonts. Text fonts need no loading at all --- set
`gp$fontfamily`, or name one for a run with `\gmfontfamily{}{}`.

## Devices and render modes

- **`"typeface"`** (default) draws glyphs as native text, so PDF and SVG
  output stays selectable and searchable: a line of prose is emitted as a
  single element, so a viewer finds a phrase and not merely a word. Text
  given a `max_width` is the exception --- it is emitted one word per
  element, because the spaces are where the lines break. Fonts are read
  straight from their OTF files,
  with no system-wide install --- but this needs a device with the R
  \eqn{\geq} 4.3 glyph engine (`ragg`, `svglite`, `cairo_pdf`). On
  others, such as the base `pdf()` device, it falls back to path mode
  with a warning.
- **`"path"`** draws each glyph as a filled vector path. Works on every
  device; the text is not selectable, and files are larger.

```r
grid.latex(r"($E = mc^2$)")                        # typeface
grid.latex(r"($E = mc^2$)", gp = gpar(fontsize = 24), render_mode = "path")  # path
```

The default devices on Windows and macOS may not find the bundled math
fonts, and warn `font family not found in Windows font database`. The
README lists the backends to prefer with the one-line setup for each,
and compares the package with `tikzDevice`, `xdvir`, `latex2exp` and
`plotmath`.

> **Do not use `showtext::showtext_auto()` with typeface mode.** showtext
> intercepts all text rendering and converts it to paths, silently
> defeating typeface mode even on `svglite` and `ragg`. Call
> `showtext::showtext_auto(FALSE)` before drawing formulas.

## Base graphics

Everything above builds grid grobs. Base graphics can render math too:
`latex_options(device_math = TRUE)` makes labels containing `$...$`
(or `\(...\)`, `\[...\]`, the same convention `latex_wrap()` uses
everywhere else) render through MicroTeX, with no other change to your plotting
code. It applies to `main`, `xlab`, `ylab`, `text()`, `mtext()`,
`legend()` and axis labels, and because the interception happens at the
device it also covers `hist()`, `boxplot()` and other packages' `plot`
methods.

### It is not limited to base graphics

The interception sits on the device, and every graphics system in R draws
its text through the same callback. So while `device_math` is on:

- `grid.text("$x^2$")` renders math.
- ggplot2 picks it up everywhere text is drawn from a string: axis
  titles, `geom_text()` and `geom_label()` labels, facet strip text,
  legend keys.
- lattice does the same for its `main`, `xlab` and panel text.

That may be exactly what you want, and it is the quickest way to get math
into a ggplot without touching the theme. Two things to weigh against it.
It is a **session-wide** switch, so a `$` in someone else's plot inside
the same script is subject to the same rule; the gate below is what
keeps that safe. And grid sizes many of its boxes from text *height*,
which is the one measurement that cannot be corrected (see
[Reserving room for a tall formula](#reserving-room-for-a-tall-formula)),
so a tall formula in a `geom_label()` or a bordered grid box can overflow
its border even though the width is right.

For ggplot2 work, `element_latex()` and `geom_latex()` remain the better
tools: they do the same rendering, scoped to the layer or theme element
you name, with no global state. Use `device_math` when you want math in
plots you are not rewriting: base plots, lattice, or a package's own
`plot()` method.

This is the whole LaTeX engine, not a translation layer, so a base plot
can carry things plotmath has no syntax for at all: a ruled table with a
coloured header row, display-style operators, calligraphic and blackboard
alphabets, automatic delimiter sizing. The coefficient table below is a
`\begin{array}` with `\hline` rules and `\rowcolor`, drawn by an ordinary
`text()` call.

```{r base-math, fig.height=4.6, fig.width=6.5}
latex_options(device_math = TRUE)

set.seed(42)
x   <- seq(0, 3, length.out = 60)
y   <- 2 + 1.4 * x^2 + rnorm(60, sd = 1.1)
fit <- lm(y ~ poly(x, 2, raw = TRUE))
cf  <- signif(coef(summary(fit)), 3)

title_eq <- r"($$\hat{y} = \beta_0 + \beta_1 x + \beta_2 x^2$$)"

# A display-style title is taller than one text line, so take the top
# margin from the measured height instead of guessing at it.
h   <- latex_dims(title_eq, gp = gpar(fontsize = par("ps")))$height
top <- ceiling(convertHeight(h, "bigpts", TRUE) /
               (par("cin")[2] * 72 * par("mex"))) + 2

par(mar = c(4.5, 5.5, top, 2))
plot(x, y, pch = 19, col = "grey45", cex = 0.7, ylim = c(0, 22),
     main = title_eq, xlab = r"($x$)",
     ylab = r"($y = f(x) + \varepsilon$)")
lines(x, predict(fit), col = "#B22222", lwd = 2)

# Keep a formula on one physical line: R hands the device a multi-line
# label one line at a time, so a table split across source lines would
# never reach it whole.
tab <- sprintf(paste0(
  r"(\begin{array}{|l|r|r|}\hline)",
  r"(\rowcolor{#E8F0FE}\text{Term} & \hat{\beta} & \text{SE}\\\hline)",
  r"(\beta_0 & %s & %s\\)",
  r"(\beta_1 & %s & %s\\)",
  r"(\beta_2 & %s & %s\\\hline\end{array})"),
  cf[1, 1], cf[1, 2], cf[2, 1], cf[2, 2], cf[3, 1], cf[3, 2])

text(0.05, 18, tab, adj = c(0, 0.5), cex = 0.75)
text(2.95, 1.2, r"($\varepsilon \sim \mathcal{N}(0, \sigma^2)$)",
     adj = c(1, 0), cex = 0.95, col = "#B22222")
```

Two placement notes, both consequences of heights being uncorrectable.
The table is anchored with `adj = c(0, 0.5)` and given headroom through
`ylim`, because R positions a top-anchored label using the *font* ascent:
a four-row table anchored at the very top of the panel would have its
header clipped away. And a formula written across several source lines
arrives at the device one line at a time and so is never recognised, which
is why `tab` is assembled with `paste0()` into a single line.

```{r base-math-off, include=FALSE}
latex_options(device_math = FALSE)
```

In an R Markdown document, set the option once in your setup chunk.
knitr captures a chunk's plots *after* its last line has run, so
switching it off at the end of the chunk that draws will leave the
figure unrendered, with the labels coming out as literal `$...$`.

### When is a label treated as math?

A dollar sign is far too common in a real axis label to treat as a
delimiter on sight: `"Cost $5-$10 per unit"` has to survive as plain
text. So a label is intercepted only when **every**
delimiter in it is closed *and* **every** `$...$` pair wraps something
that actually looks like math (a command, a script, or a lone
variable):

| Label | Rendered as | Why |
|:--|:--|:--|
| `"Revenue ($)"` | text | one unclosed `$` |
| `"Price is $5 today"` | text | one unclosed `$` |
| `"Cost $5-$10"` | text | closed, but `5-` is not math |
| `"Budget $1,000 to $5,000 with $x$ shown"` | text | `1,000 to ` is not math |
| `"$x$"` | math | a lone variable |
| `"$x^2$"` | math | a script |
| `"Slope $\\hat{\\beta}_1$"` | math | a command |
| `"\\(\\alpha\\)"` | math | explicit delimiters |

Every pair has to qualify, not merely one of them: `latex_wrap()` cannot
be told to treat one span as math and leave another alone, so a single
good `$x$` would otherwise drag the currency in with it.

Everything that fails the test is handed to the device untouched, byte
for byte. Turning the option on cannot change how an existing plot
looks unless the label really does contain math. Write `\$` for a
literal dollar next to real math in the same label.

### Reserving room for a tall formula

Widths are exact while the option is on: `strwidth()` reports the
formula's real width, so labels centre correctly. **Heights are not, and
cannot be.** R computes text height from the font, never from the
string: `strheight()` returns the same value for `"abc"` and for
`$\sum_{i=1}^{n}\frac{x_i}{2}$`, and a graphics device has no
string-height entry point to intercept. A tall formula can therefore
overflow a `legend()` box or the margin `par("mar")` set aside for it.

Measure it with `latex_dims()` and reserve the room yourself:

```{r base-margin}
h  <- latex_dims(r"($\frac{a}{b}$)", gp = gpar(fontsize = par("ps")))$height
bp <- convertHeight(h, "bigpts", TRUE)

# par(mar) counts lines of par("cin"), so convert in those units rather
# than grid's "lines" -- the two agree only while gpar$fontsize happens
# to equal par("ps").
bp / (par("cin")[2] * 72 * par("mex"))
```

Round that up and add it to the relevant side of `par(mar = )`.

The other differences from the grid path: math is drawn as vector
outlines, so unlike `latex_grob()` it is not selectable in a PDF; prose
in a bold label is not bolded, since face comes from `\textbf` rather
than from the device; `\includegraphics` is left out and rounded box
corners are squared. And `expression()` labels are not affected at all: R lays
plotmath out inside the graphics engine, where a device never sees it.
`?latex_options` lists the rest.

## Utilities

### Measuring

`latex_dims()` returns the bounding box of an expression, for layout
arithmetic and for checking that a label fits:

```{r dims}
latex_dims(r"(\frac{a}{b})", gp = gpar(fontsize = 20))
```

### Session defaults

`latex_options()` sets `math_font`, `render_mode` and `input_mode` for
calls that don't supply them; explicit arguments always win. Size stays at
the grob level, via `gp$fontsize` / `gp$lineheight`.

```{r options, eval=FALSE}
latex_options(math_font = "stix", render_mode = "typeface")
latex_options()        # query
reset_latex_options()  # back to built-in defaults
```

### User-defined macros

`define_macro()` registers zero-argument shorthands, expanded by text
substitution before the expression reaches MicroTeX:

```{r macros, fig.height=0.7, fig.width=3, out.width="50%"}
define_macro("RR", r"(\mathbb{R})")
define_macro("eps", r"(\varepsilon)")

grid.newpage()
grid.latex(r"(\forall \eps > 0, \eps \in \RR)")

clear_macros()
```

Names must be ASCII letters, and expansion iterates to a fixed point so
macros can reference each other. `list_macros()` shows what is registered;
`clear_macros()` drops everything.

For parameterised macros (0--9 arguments) MicroTeX also accepts plain-TeX
`\def`. These live only for the expression they appear in, so use them for
an abbreviation local to one label and `define_macro()` for one that
should persist:

```{r def-inline, fig.height=0.7, fig.width=4, out.width="60%"}
grid.newpage()
grid.latex(
  r"(\def\norm#1{\left\lVert #1 \right\rVert}
      \norm{\vec{v}} = \sqrt{\langle \vec{v}, \vec{v} \rangle})"
)
```

### Caching and introspection

Parsed layouts are memoised by `(tex, fontsize, math_font, render_mode,
…)`, so a repeated axis label is laid out once:

```{r cache, eval=FALSE}
latex_cache_info()       # size / max_size / hits / misses
latex_cache_limit(1024)  # LRU capacity; 0 disables caching
latex_cache_clear()      # wipe (e.g. after re-loading fonts)
```

`latex_tree()` returns the raw draw records plus bbox metadata, and
`debug = TRUE` overlays the bounding box, baseline and record origins ---
both useful when checking alignment:

```{r debug, fig.height=1, fig.width=3, out.width="60%"}
grid.newpage()
grid.latex(r"($x^{2} + y_{i}$)", debug = TRUE)
```

## LaTeX reference

MicroTeX is a **math formula renderer**, not a document typesetter. It
covers the vast majority of notation used in plots and figures, but does
not replace a LaTeX installation. This section is the boundary.

### Lists

`itemize` and `enumerate` lay their items out as a left-aligned column,
one per row --- `itemize` prefixes a bullet, `enumerate` numbers them:

```{r lists, fig.height=1.4, fig.width=3, out.width="45%"}
grid.newpage()
grid.latex(r"(\begin{enumerate}
  \item e^{i\pi} + 1 = 0
  \item \begin{itemize}
          \item \alpha \item \beta
        \end{itemize}
\end{enumerate})", gp = gpar(fontsize = 20))
```

An optional `[…]` argument customises the marker. For `itemize` it is the
literal marker (`\begin{itemize}[\star]`); for `enumerate` it is a counter
template containing one of `\arabic*`, `\alph*`, `\Alph*`, `\roman*` or
`\Roman*` (e.g. `\begin{enumerate}[\Roman*.]`). Lists nest, and an item
may contain any math, including a `\begin{array}` table.

Because MicroTeX is a math engine, each item is a **math-mode,
single-line** expression: no paragraph flow, no line wrapping, and prose
inside an item needs `\text{}` (`\item \text{First point}`). The
`description` environment is not supported.

### Pasting LaTeX from other sources

Input generated by other tools --- ready-to-compile `tabular` snippets,
fragments copied out of a `.tex` file --- usually arrives wrapped in
document-level constructs MicroTeX does not implement. Rather than
refusing it, gridmicrotex rewrites or removes a small set of well-known
wrappers before parsing, so `knitr::kable(format = "latex")` and
`xtable::print.xtable()` output can be pasted in unedited:

```{r pasted, fig.height = 1.1, fig.width = 3.2, out.width = "55%"}
snippet <- r"(
% latex table generated by kable()
\begin{table}[ht]
\centering
\caption{Model coefficients}
\begin{tabular}{lrr}
\toprule
Term & Estimate & \emph{p} \\
\midrule
Intercept & 2.14 & 0.003 \\
Slope & 0.42 & 0.001 \\
\bottomrule
\end{tabular}
\end{table}
)"

grid.newpage()
grid.latex(snippet, input_mode = "mixed", gp = gpar(fontsize = 11))
```

Six things were handled without any editing: the `%` comment and the
`table` float were dropped, `\centering` removed, `\caption` set as a line
above the table, `\toprule`/`\bottomrule` became thick rules, `\midrule` a
plain one, and `\emph` became italic.

Note that the cell text is set in math italics. A `tabular` is a math
environment, so its contents are math whichever `input_mode` you choose ---
the two modes give an identical layout for the snippet above. What
`"mixed"` buys you here is the prose *outside* the environment, such as the
caption. For upright cell text, wrap the cells in `\text{}`.

**Removed silently (no visual effect):**

| Construct | Why |
|---|---|
| `%`-to-end-of-line comments (`\%` is preserved) | comments are non-visual in LaTeX too |
| `\documentclass[…]{…}`, `\usepackage[…]{…}` | preamble metadata |
| `\begin{document}` / `\end{document}` | document boundary, structural only |
| `\maketitle`, `\title{…}`, `\author{…}` | title-page metadata, no body output |
| `\label{…}` | cross-reference target, never rendered in LaTeX either |
| `\begin{table}[…]` / `\end{table}`, `\begin{figure}[…]` / `\end{figure}` (and starred variants) | float wrappers; the contents stay |
| `\centering`, `\raggedright`, `\raggedleft`, `\flushleft`, `\flushright` | alignment scope declarations |
| `\noindent`, `\relax` | content-free declarations |

**Rewritten to a MicroTeX equivalent:**

| Construct | Becomes |
|---|---|
| `\emph{X}` | `\textit{X}` |
| `\textnormal{X}` | `\text{X}` |
| `\par`, `\newline` | `\\` (line break) |
| `\toprule`, `\bottomrule` | `\thickhline` (rendered ~2× thickness) |
| `\midrule` | `\hline` |
| `\cmidrule[trim]?(parenarg)?{a-b}` | `\cline{a-b}` --- partial-column rule |
| `\caption[short]{X}` | `\text{X}` plus a line break, at its source position --- so a caption written above the table stays above it |
| `\smallskip`, `\medskip`, `\bigskip` | `\vspace{0.25em}` / `\vspace{0.5em}` / `\vspace{1em}` --- em-relative so they scale with `gp$fontsize` |
| `\hfill`, `\vfill` | `\quad` / `\vspace{1em}` --- static proxies for rubber lengths |
| `\url{X}` | coloured monospace text --- a grob cannot be a hyperlink, so only the appearance survives |
| `\href{U}{X}` | `X`, coloured. Matches `hyperref` with `colorlinks=true`; markdown links use HTML's blue-and-underlined convention instead |

Three of those are approximations. `\caption` renders where it appears in
the source, not where LaTeX's float machinery would move it, so
caption-above or caption-below follows whatever your tool emits. The skips
are em-relative rather than LaTeX's absolute 3/6/12 pt, so they stay
visible at any `gp$fontsize`; use `\vspace{Xpt}` for an exact amount. And
`\hfill` / `\vfill` are *rubber* lengths with nothing to fill in a
fixed-size grob, so they become a static 1 em gap --- right position, no
elasticity.

**Not honored** --- rendered as literal text, which is intentional: it
makes unsupported markup easy to spot.

- Declarative font scopes: `\bfseries`, `\itshape`, `\ttfamily`,
  `\sffamily`, `\rmfamily`. These affect text within their group in LaTeX,
  which needs scope tracking we do not implement. Use the
  argument-bearing forms instead --- `\textbf{…}`, `\textit{…}`,
  `\texttt{…}`, `\textsf{…}`, `\textrm{…}` --- all of which nest. To
  choose the text font itself, set `gp$fontfamily` or use
  `\gmfontfamily{…}{…}`.
- References: `\ref{…}`, `\cite{…}` --- there is nothing to resolve
  against.
- Footnotes: `\footnote{…}` --- the positioning machinery is page-bound.
- Small caps: `\textsc{…}` --- MicroTeX has no small-caps glyphs.

### What is not supported

These need a real document compiler and are outside a formula renderer's
scope: document structure (`\section`, page layout, `\tableofcontents`);
automatic hyphenation (`\-` marks a break point yourself, and
`max_width` / `justify` do handle line breaking); TikZ/PGF;
cross-references and bibliographies; theorem
environments; the `description` list environment (`itemize` and
`enumerate` *are* supported); and `\tag` / equation numbering.
`\usepackage{…}` is accepted but loads nothing --- every supported command
is built into MicroTeX.

For axis labels, annotations, legends and in-plot formulas, the supported
set is more than sufficient.
