---
title: "Using LaTeX Math in ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using LaTeX Math in ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 300,
  dev = "ragg_png"
)
library(gridmicrotex)
library(ggplot2)
```

gridmicrotex provides ggplot2 extensions for rendering LaTeX math in
plots:

- **`geom_latex()`** --- a geom layer for placing LaTeX labels at data
  coordinates.
- **`element_latex()`** --- a theme element for rendering axis titles,
  plot titles, and other text elements as LaTeX.

There are matching `geom_markdown()` and `element_markdown()` for labels
written in markdown rather than raw LaTeX --- covered at the end of this
vignette, and in more depth in `vignette("markdown")`.

## Annotating plots with `geom_latex()`

`geom_latex()` works like `geom_text()` but interprets the `label`
aesthetic as a LaTeX math string. You can also map the `size` (font size in points) 
and `colour` aesthetics as usual. `element_latex()` replaces a text theme element so that its label is
rendered as LaTeX math.

```{r geom-basic, out.width="70%"}
df <- data.frame(
  x = 1:3,
  y = 1:3,
  eq = c("$x^2$", "\\frac{a}{b}", "$\\sum_{i=1}^n x_i$"),
  col = c("red", "blue", "green")
)

ggplot(df, aes(x, y, 
               label = eq, 
               colour = col, 
               size = c(14, 18, 14))) +
  geom_latex() +
  scale_colour_identity() +
  scale_size_identity() +
  labs(
    x = "$\\beta_1 \\cdot x + \\beta_0$",
    y = "$\\mathrm{mpg}$"
  ) +
  theme(
    axis.title.x = element_latex(fontsize = 14),
    axis.title.y = element_latex(fontsize = 14)
  )
```

Dollar-sign delimiters (`$...$`) are stripped automatically, so both
`"\\frac{a}{b}"` and `"$\\frac{a}{b}$"` produce the same output.


### Adding equation annotations to a scatter plot

A common use case is annotating a regression fit with the model
equation. Use `annotate("latex", ...)` for single annotations --- it
delegates to `GeomLatex` internally but avoids creating a data frame
and automatically hides the legend.

```{r regression-annotation, out.width="70%"}
fit <- lm(mpg ~ wt, data = mtcars)
b0 <- round(coef(fit)[1], 1)
b1 <- round(coef(fit)[2], 1)
r2 <- round(summary(fit)$r.squared, 3)

eq_label <- sprintf("$\\hat{y} = %s %s x, \\quad R^2 = %s$",
                     b0, b1, r2)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("latex", x = 4, y = 30, label = eq_label, size = 12) +
  theme_minimal()
```

## Markdown labels

When a label is more prose than formula --- a bold phrase, an italic
word, a symbol in a sentence --- `geom_markdown()` and
`element_markdown()` accept markdown with inline `$math$`. They take the
same aesthetics and theme slots as their LaTeX counterparts; only the
label syntax differs.

```{r markdown, out.width="70%"}
df <- data.frame(
  x   = 1:3,
  y   = c(2, 3, 1),
  lab = c("**bold**", "*slope* $\\beta_1$", "`code` and $x^2$")
)

ggplot(df, aes(x, y, label = lab)) +
  geom_point() +
  geom_markdown(fontsize = 12, vjust = -0.6) +
  ylim(0.5, 3.6) +
  labs(
    title = "*Fitted* model: $\\hat{y} = \\beta_0 + \\beta_1 x$",
    x     = "**weight** in $10^3$ lbs",
    y     = "*efficiency* $\\eta$"
  ) +
  theme(
    plot.title   = element_markdown(fontsize = 14),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
```

Unlike `element_latex()`, `element_markdown()` never strips `$`
delimiters --- in markdown a `$...$` pair *is* the math, so removing it
would change the label. Note that the [ggtext](https://wilkelab.org/ggtext/)
package also exports an `element_markdown()`; if both are attached, call
`gridmicrotex::element_markdown()` to be explicit. The two are not
interchangeable --- ggtext renders HTML/CSS, this renders LaTeX math.

### Annotating with markdown

`annotate("markdown", ...)` is the markdown counterpart of
`annotate("latex", ...)`. It suits a callout that is part prose and part
formula --- and because `<br>` starts a new line, one annotation can hold
several:

```{r markdown-annotation, out.width="70%"}
fit <- lm(mpg ~ wt, data = mtcars)

note <- sprintf(
  "**Linear fit**<br>$\\hat{y} = %s %s x$<br>$R^2 = %s$, *p* < 0.001",
  round(coef(fit)[1], 1),
  round(coef(fit)[2], 1),
  round(summary(fit)$r.squared, 3)
)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point(colour = "grey65") +
  geom_smooth(method = "lm", se = FALSE, colour = "#1F6FB2") +
  annotate("markdown", x = 4.1, y = 32, label = note, size = 11,
           style = "strong { color: #B22222 }") +
  theme_minimal()
```

The LaTeX version of the same annotation would need `\text{}` around
every word and could not set the heading in bold on its own line. As
with `annotate("latex", ...)`, no data frame is needed and the legend is
suppressed automatically.

### Styling labels

Both take a `style`: a `markdown_style()` object, CSS text, or the path
to a `.css` file. It is the same cascade `markdown_box_grob()` uses --- so
a stylesheet can set the look of every label at once, and `<span
class=>` picks out one run inside a label.

```{r style, out.width="70%"}
css <- "
  body   { color: #33475B }
  strong { color: #B22222 }
  code   { color: #1F6FB2 }
  .unit  { color: grey55; font-size: smaller }
"

df <- data.frame(
  x = 1:3, y = c(2, 3, 1),
  lab = c("**bold** is red", "`code` is blue",
          'plain <span class="unit">with a note</span>')
)

ggplot(df, aes(x, y, label = lab)) +
  geom_point() +
  geom_markdown(style = css, fontsize = 11, vjust = -0.8) +
  ylim(0.5, 3.8) +
  labs(title = "**Styled** labels and $\\beta_1$",
       x = 'weight <span class="unit">(1000 lbs)</span>') +
  theme(
    plot.title   = element_markdown(style = css, fontsize = 14),
    axis.title.x = element_markdown(style = css)
  )
```

Only the properties that compile to LaTeX apply --- colour, the `font-*`
family, decorations. A label has no block layout, so margins, padding and
backgrounds are ignored; `markdown_box_grob()` is the place for those.
`latex_options(markdown_style = )` sets a default for a whole document,
so the argument is only needed to override it.





