---
title: "Using LaTeX Math in ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using LaTeX Math in ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 300
)
library(gridmicrotex)
library(ggplot2)
```

gridmicrotex provides two ggplot2 extensions for rendering LaTeX math
in plots:

- **`geom_latex()`** --- a geom layer for placing LaTeX labels at data
  coordinates.
- **`element_latex()`** --- a theme element for rendering axis titles,
  plot titles, and other text elements as LaTeX.

## Annotating plots with `geom_latex()`

`geom_latex()` works like `geom_text()` but interprets the `label`
aesthetic as a LaTeX math string.

```{r geom-basic, out.width="70%"}
df <- data.frame(
  x = 1:3,
  y = 1:3,
  eq = c("x^2", "\\frac{a}{b}", "\\sum_{i=1}^n x_i")
)

ggplot(df, aes(x, y, label = eq)) +
  geom_latex() +
  theme_minimal()
```

### Controlling size and colour

Map the `size` (font size in points) and `colour` aesthetics as usual:

```{r geom-aesthetics, out.width="70%"}
df$col <- c("red", "blue", "darkgreen")

ggplot(df, aes(x, y, label = eq, colour = col, size = c(14, 18, 14))) +
  geom_latex() +
  scale_colour_identity() +
  scale_size_identity() +
  theme_minimal()
```

### Adding equation annotations to a scatter plot

A common use case is annotating a regression fit with the model
equation. Use `annotate("latex", ...)` for single annotations --- it
delegates to `GeomLatex` internally but avoids creating a data frame
and automatically hides the legend.

```{r regression-annotation, out.width="70%"}
fit <- lm(mpg ~ wt, data = mtcars)
b0 <- round(coef(fit)[1], 1)
b1 <- round(coef(fit)[2], 1)
r2 <- round(summary(fit)$r.squared, 3)

eq_label <- sprintf("\\hat{y} = %s %s x, \\quad R^2 = %s",
                     b0, b1, r2)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("latex", x = 4, y = 30, label = eq_label, size = 12) +
  theme_minimal()
```

## LaTeX axis titles with `element_latex()`

`element_latex()` replaces a text theme element so that its label is
rendered as LaTeX math.

```{r element-axis, out.width="70%"}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(
    x = "$\\beta_1 \\cdot x + \\beta_0$",
    y = "$\\mathrm{mpg}$"
  ) +
  theme(
    axis.title.x = element_latex(fontsize = 14),
    axis.title.y = element_latex(fontsize = 14)
  )
```

Dollar-sign delimiters (`$...$`) are stripped automatically, so both
`"\\frac{a}{b}"` and `"$\\frac{a}{b}$"` produce the same output.

