---
title: "Introduction to gridmicrotex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gridmicrotex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "ragg_png"
)
library(gridmicrotex)
```

## What is gridmicrotex?

**gridmicrotex** renders LaTeX math equations as native R `grid` graphics
objects (grobs). It uses the [MicroTeX](https://github.com/NanoMichael/MicroTeX)
C++ library as its layout engine --- MicroTeX parses LaTeX, builds the TeX
box model, and computes exact glyph coordinates. The package intercepts
this layout data and maps it to native grid primitives (`pathGrob`,
`segmentsGrob`, `rectGrob`, `textGrob`), producing a `gTree` that works
on any R graphics device at any resolution.

**Key features:**

- No external LaTeX installation required --- MicroTeX is fully embedded
- Resolution-independent vector output on all R devices (PNG, PDF, SVG, ...)
- Full math support: fractions, roots, integrals, matrices, Greek letters,
  accents, delimiters, and more
- Two bundled math fonts: Lete Sans Math (sans-serif, default) and
  STIX Two Math (serif); additional fonts via `load_font()`
- Color support via `\textcolor{}`
- ggplot2 integration with `geom_latex()` and `element_latex()`
- CJK and multilingual text in `\text{}` blocks

## Basic usage

The core function is `latex_grob()`, which returns a grid grob:

```{r basic, fig.height=0.8, fig.width=2, out.width="50%"}
g <- latex_grob("\\frac{-b \\pm \\sqrt{b^2 - 4ac}}{2a}", gp = grid::gpar(fontsize = 24))
grid::grid.newpage()
grid::grid.draw(g)
```

For quick rendering, use `grid.latex()`:

```{r quick, fig.height=0.8, fig.width=2, out.width="50%"}
grid::grid.newpage()
grid.latex("\\sum_{i=1}^{n} x_i^2", gp = grid::gpar(fontsize = 28))
```

## Positioning and justification

Control placement with `x`, `y`, `hjust`, and `vjust`:

```{r positioning, fig.height=1.2, fig.width=2, out.width="50%"}
grid::grid.newpage()
grid.latex("E = mc^2", x = 0.2, y = 0.7, hjust = 0, gp = grid::gpar(fontsize = 24))
grid.latex("F = ma", x = 0.2, y = 0.3, hjust = 0, gp = grid::gpar(fontsize = 24))
```

## Colors

Set the formula color via `gp`, or use `\textcolor{}` within the LaTeX:

```{r colors, fig.height=0.6, fig.width=2, out.width="40%"}
grid::grid.newpage()
grid.latex(
  "\\textcolor{red}{\\alpha} + \\textcolor{blue}{\\beta} = \\gamma",
  gp = grid::gpar(fontsize = 28)
)
```

## Math fonts

The package ships with two math fonts, both loaded automatically:

| Alias | Font | License |
|-------|------|---------|
| `"lete"` (default) | Lete Sans Math | SIL Open Font License |
| `"stix"` | STIX Two Math | SIL Open Font License |

```{r fonts-list}
available_math_fonts()
```

```{r fonts-default, fig.height=0.8, fig.width=2, out.width="50%"}
latex_options(math_font = "stix")
grid::grid.newpage()
grid.latex("\\int_0^1 f(x)\\,dx", gp = grid::gpar(fontsize = 24))

# Switch back to the default (Lete Sans Math)
latex_options(math_font = "lete")
```

You can also override the font per call via `math_font`:

```{r fonts, fig.height=1.5, fig.width=2, out.width="50%"}
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(2, 1)))
grid::pushViewport(grid::viewport(layout.pos.row = 1))
grid.latex("\\int_0^1 f(x)\\,dx", gp = grid::gpar(fontsize = 24))
grid::upViewport()
grid::pushViewport(grid::viewport(layout.pos.row = 2))
grid.latex("\\int_0^1 f(x)\\,dx", gp = grid::gpar(fontsize = 24), math_font = "stix")
grid::upViewport(2)
```

Use `available_math_fonts()` to list loaded fonts and `check_fonts()`
for a diagnostic report.

### Advanced: loading custom fonts

Use `load_font()` to add any additional OpenType math font. The OpenType
MATH table is parsed directly in C++ --- no companion metrics file or
external toolchain is required:

```r
load_font("path/to/MyFont.otf")
```

### Render modes

gridmicrotex supports two rendering modes for math glyphs:
  
- **`"typeface"`** (default): Renders glyphs as native text using the
  math font's typeface. This produces selectable, searchable, and
  accessible text in PDF and SVG output. Requires the math font (e.g.,
  Lete Sans Math) to be installed on the system, and a device that
  supports font embedding (e.g., `ragg::agg_png()`,
  `svglite::svglite()`, `grDevices::cairo_pdf()`). On devices that
  do not support typeface rendering (e.g., the base `pdf()` device),
  the package automatically falls back to path mode with a warning.

- **`"path"`**: Renders each glyph as a filled vector path.
  This works on all R graphics devices and produces pixel-perfect output.
  However, text in PDF/SVG output is not selectable or searchable.

```r
# Default typeface mode (selectable text in PDF/SVG)
grid.latex("E = mc^2", gp = grid::gpar(fontsize = 24))

# Explicit path mode (works everywhere, but text is not selectable)
grid.latex("E = mc^2", gp = grid::gpar(fontsize = 24), render_mode = "path")
```

> **Important: Do not use `showtext::showtext_auto()` with typeface mode.**
> The [showtext](https://CRAN.R-project.org/package=showtext) package
> globally intercepts all text rendering and converts it to vector paths.
> This silently defeats typeface mode, causing all math glyphs to appear
> as paths instead of native text — even on devices like `svglite` and
> `ragg` that fully support font embedding. If you need showtext for
> other parts of your plot, disable it before drawing LaTeX formulas:
>
> ```r
> showtext::showtext_auto(FALSE)
> grid.latex("E = mc^2", gp = grid::gpar(fontsize = 24))  # typeface mode works correctly
> ```

## Querying dimensions

`latex_dims()` returns the bounding box of an expression:

```{r dims}
dims <- latex_dims("\\frac{a}{b}", gp = grid::gpar(fontsize = 20))
dims
```

This is useful for layout calculations and ensuring labels fit.

## Text rendering and CJK support

Text inside `\text{}` and `\mbox{}` is rendered using R's standard
text-rendering system. This means `gp$fontfamily` controls the font for
**all** text content --- Latin letters, CJK characters, Cyrillic, and any
other script your R graphics device supports:

```{r cjk, fig.height = 0.8, fig.width = 2.5, out.width="50%"}
grid::grid.newpage()
grid.latex("x^2 + \\text{你好}", gp = grid::gpar(fontsize = 24, fontfamily = "sans"))
```

Any font available to R works: base families like `"sans"`, `"serif"`,
`"mono"`, or fonts registered via **showtext** / **systemfonts**.

### Font pairing

The bundled math fonts have different styles. For a consistent look,
pair them with a matching `fontfamily`:

| Math font | Style | Suggested `fontfamily` |
|-----------|-------|------------------------|
| Lete Sans Math (`"lete"`, default) | Sans-serif | `"sans"` |
| STIX Two Math (`"stix"`) | Serif | `"serif"` |

```{r font-pairing, fig.height = 1, fig.width = 6, out.width="70%"}
grid::grid.newpage()
grid.latex(
  "\\text{Theorem: } \\forall x \\in \\mathbb{R},\\; x^2 \\geq 0",
  math_font = "stix",
  gp = grid::gpar(fontfamily = "serif", fontsize = 12)
)
```

## Supported LaTeX

gridmicrotex uses the MicroTeX engine, which is a **math formula renderer**,
not a full document typesetter. It covers the vast majority of math notation
you would use in plots and figures, but does not attempt to replace a full
LaTeX installation.

### Complicated examples

```{r complex-formula, fig.height = 3, fig.width=6, out.width="60%"}
grid::grid.newpage()
grid.latex(paste0(
      "\\begin{array}{l}",
      "  \\forall\\varepsilon\\in\\mathbb{R}_+^*\\ \\exists\\eta>0",
      "\\ |x-x_0|\\leq\\eta\\Longrightarrow|f(x)-f(x_0)|\\leq\\varepsilon\\\\",
      "  \\det",
      "  \\begin{bmatrix}",
      "      a_{11}&a_{12}&\\cdots&a_{1n}\\\\",
      "      a_{21}&\\ddots&&\\vdots\\\\",
      "      \\vdots&&\\ddots&\\vdots\\\\",
      "      a_{n1}&\\cdots&\\cdots&a_{nn}",
      "  \\end{bmatrix}",
      "  \\overset{\\mathrm{def}}{=}\\sum_{\\sigma\\in\\mathfrak{S}_n}",
      "\\varepsilon(\\sigma)\\prod_{k=1}^n a_{k\\sigma(k)}\\\\",
      "  \\int_0^\\infty{x^{2n} e^{-a x^2}\\,dx} = \\frac{2n-1}{2a}",
      " \\int_0^\\infty{x^{2(n-1)} e^{-a x^2}\\,dx}",
      " = \\frac{(2n-1)!!}{2^{n+1}} \\sqrt{\\frac{\\pi}{a^{2n+1}}}\\\\",
      "\\end{array}"
), gp = grid::gpar(fontsize = 16))
```


```{r table-multicolumn, fig.height = 3, fig.width=8, out.width="60%"}
grid::grid.newpage()

grid.latex(
  "
  \\newcolumntype{s}{>{\\color{#1234B6}}c}
\\begin{array}{|c|c|c|s|}
  \\hline
  \\rowcolor{Tan}\\multicolumn{4}{|c|}{\\textcolor{white}{\\bold{\\text{Table Head}}}}\\\\
  \\hline
  \\text{Matrix}&\\multicolumn{2}{|c|}{\\text{Multicolumns}}&\\text{Font size commands}\\\\
  \\hline
  \\begin{pmatrix}
      \\alpha_{11}&\\cdots&\\alpha_{1n}\\\\
      \\hdotsfor{3}\\\\
      \\alpha_{n1}&\\cdots&\\alpha_{nn}
  \\end{pmatrix}
  &\\large \\text{Left}&\\cellcolor{#00bde5}\\small \\textcolor{white}{\\text{\\bold{Right}}}
  &\\small \\text{small Small}\\\\
  \\hline
  \\multicolumn{4}{|c|}{\\text{Table Foot}}\\\\
  \\hline
\\end{array}
  ",
  gp = grid::gpar(fontsize = 22)
)
```

```{r complicated-equation, fig.height = 9, fig.width=9, out.width="60%"}
grid::grid.newpage()
grid.latex(
  "\\definecolor{gris}{gray}{0.9}
\\definecolor{noir}{rgb}{0,0,0}
\\definecolor{bleu}{rgb}{0,0,1}
\\fatalIfCmdConflict{false}
\\newcommand{\\pa}{\\left|}
\\begin{array}{c}
  \\LaTeX\\\\
  \\begin{split}
      |I_2| &= \\pa\\int_0^T\\psi(t)\\left\\{ u(a,t)-\\int_{\\gamma(t)}^a \\frac{d\\theta}{k} (\\theta,t) \\int_a^\\theta c(\\xi)
          u_t (\\xi,t)\\,d\\xi\\right\\}dt\\right|\\\\
      &\\le C_6 \\Bigg|\\pa f \\int_\\Omega \\pa\\widetilde{S}^{-1,0}_{a,-}
          W_2(\\Omega, \\Gamma_1)\\right|\\ \\right|\\left| |u|\\overset{\\circ}{\\to} W_2^{\\widetilde{A}}(\\Omega\\Gamma_r,T)\\right|\\Bigg|\\\\
      &\\\\
      &\\begin{pmatrix}
          \\alpha&\\beta&\\gamma&\\delta\\\\
          \\aleph&\\beth&\\gimel&\\daleth\\\\
          \\mathfrak{A}&\\mathfrak{B}&\\mathfrak{C}&\\mathfrak{D}\\\\
          \\boldsymbol{\\mathfrak{a}}&\\boldsymbol{\\mathfrak{b}}&\\boldsymbol{\\mathfrak{c}}&\\boldsymbol{\\mathfrak{d}}
      \\end{pmatrix}
      \\quad{(a+b)}^{\\frac{n}{2}}=\\sqrt{\\sum_{k=0}^n\\tbinom{n}{k}a^kb^{n-k}}\\quad
          \\Biggl(\\biggl(\\Bigl(\\bigl(()\\bigr)\\Bigr)\\biggr)\\Biggr)\\\\
      &\\forall\\varepsilon\\in\\mathbb{R}_+^*\\ \\exists\\eta>0\\ |x-x_0|\\leq\\eta\\Longrightarrow|f(x)-f(x_0)|\\leq\\varepsilon\\\\
      &\\det
      \\begin{bmatrix}
          a_{11}&a_{12}&\\cdots&a_{1n}\\\\
          a_{21}&\\ddots&&\\vdots\\\\
          \\vdots&&\\ddots&\\vdots\\\\
          a_{n1}&\\cdots&\\cdots&a_{nn}
      \\end{bmatrix}
      \\overset{\\mathrm{def}}{=}\\sum_{\\sigma\\in\\mathfrak{S}_n}\\varepsilon(\\sigma)\\prod_{k=1}^n a_{k\\sigma(k)}\\\\
      &\\Delta f(x,y)=\\frac{\\partial^2f}{\\partial x^2}+\\frac{\\partial^2f}{\\partial y^2}\\qquad\\qquad \\fcolorbox{noir}{gris}
          {n!\\underset{n\\rightarrow+\\infty}{\\sim} {\\left(\\frac{n}{e}\\right)}^n\\sqrt{2\\pi n}}\\\\
      &\\sideset{_\\alpha^\\beta}{_\\gamma^\\delta}{
      \\begin{pmatrix}
          a&b\\\\
          c&d
      \\end{pmatrix}}
      \\xrightarrow[T]{n\\pm i-j}\\sideset{^t}{}A\\xleftarrow{\\overrightarrow{u}\\wedge\\overrightarrow{v}}
          \\underleftrightarrow{\\iint_{\\mathds{R}^2}e^{-\\left(x^2+y^2\\right)}\\,\\mathrm{d}x\\mathrm{d}y}
  \\end{split}\\\\
  \\rotatebox{30}{\\sum_{n=1}^{+\\infty}}\\quad\\mbox{Mirror rorriM}\\reflectbox{\\mbox{Mirror rorriM}}
\\end{array}",
  gp = grid::gpar(fontsize = 22),
  render_mode = "path"
)
```



### What is not supported

MicroTeX is a **math formula renderer**, not a full LaTeX engine. The
following are outside its scope:

- **Document structure**: `\section`, `\begin{document}`, page layout,
  headers/footers, `\tableofcontents`
- **Package loading**: `\usepackage{}` --- all supported commands are
  built in
- **Paragraph text**: line breaking, hyphenation, justified paragraphs
- **TikZ / PGF** drawing commands
- **Images**: `\includegraphics`
- **Cross-references**: `\label`, `\ref`, `\cite`, bibliographies
- **Theorem environments**: `\begin{theorem}`, `\begin{proof}`
- **Lists**: `itemize`, `enumerate`, `description`
- **Some amsmath commands**: `\substack`, `\tag`, equation numbering

For most statistical graphics use cases --- axis labels, annotations,
legends, and in-plot formulas --- the supported feature set is more
than sufficient.

## Project-wide defaults

`latex_options()` sets defaults for `math_font` and `render_mode`, used
by `latex_grob()`, `grid.latex()`, `latex_dims()`, and `latex_tree()`
whenever the corresponding argument is not supplied at the call site.
Size is controlled at the grob level via `gp$fontsize` /
`gp$lineheight` (see *Basic usage*).

```{r options, eval=FALSE}
latex_options(math_font = "stix", render_mode = "typeface")

# Later calls pick these up automatically
grid.latex("\\sum_{i=1}^{n} i^{2}", gp = grid::gpar(fontsize = 14))

# Query current settings
latex_options()

# Reset to built-in defaults
reset_latex_options()
```

Explicit arguments always win. Setting `math_font` via
`latex_options()` also updates the MicroTeX engine default, so you
don't also need a separate font-setup call.

## User-defined macros

`define_macro()` registers zero-argument shorthands that are expanded
by text substitution before the expression reaches MicroTeX. Handy for
recurring notation:

```{r macros, fig.height=0.7, fig.width=3, out.width="50%"}
define_macro("RR", "\\mathbb{R}")
define_macro("eps", "\\varepsilon")

grid::grid.newpage()
grid.latex("\\forall \\eps > 0, \\eps \\in \\RR", gp = grid::gpar(fontsize = 24))

clear_macros()
```

Macro names must be ASCII letters. Expansion iterates to a fixed point,
so macros can reference other macros. Use `list_macros()` to see
currently registered ones, and `clear_macros()` (with no arguments) to
drop them all.

## Layout caching

Parsed layouts are memoised by `(tex, fontsize, math_font, render_mode,
...)`. Re-drawing the same formula --- for example, the same axis label
across many plots --- reuses the cached layout:

```{r cache, eval=FALSE}
latex_cache_info()       # size / max_size / hits / misses
latex_cache_limit(1024)  # raise or lower the LRU capacity
latex_cache_clear()      # wipe the cache (e.g. after re-loading fonts)
```

Set the limit to `0` to disable caching entirely.

## Introspecting a formula

`latex_tree()` returns the raw draw-record table plus bbox metadata,
useful for debugging alignment, counting glyphs, or building custom
grobs on top of the layout:

```{r tree}
tr <- latex_tree("\\frac{a}{b}")
tr
head(tr$records, 3)
```

## Debug overlay

Pass `debug = TRUE` to `latex_grob()` / `grid.latex()` to overlay
diagnostics on the rendered formula --- the full bounding box (dashed
gray), the baseline (solid red), and a dot at each draw record's
origin. Useful for checking vertical alignment between a formula and
surrounding grobs:

```{r debug, fig.height=1, fig.width=3, out.width="60%"}
grid::grid.newpage()
grid.latex("x^{2} + y_{i}", gp = grid::gpar(fontsize = 30), debug = TRUE)
```

## Comparison with alternatives

| Approach       | LaTeX required? | Device independent? | Vector? | Math coverage |
|:---------------|:---------------:|:-------------------:|:-------:|:-------------:|
| `tikzDevice`   | Yes             | No                  | Yes     | Full          |
| `xdvir`        | Yes             | No                  | Yes     | Full          |
| `latexpdf`     | Yes             | No                  | Yes     | Full (tables) |
| `latex2exp`    | No              | Yes                 | Yes     | Limited       |
| `plotmath`     | No              | Yes                 | Yes     | Limited       |
| **gridmicrotex** | **No**       | **Yes**             | **Yes** | **Broad**     |


## Graphics backend

The default graphics device on Windows (`windows()`) and macOS
(`quartz()`) may not find the bundled math fonts, producing warnings
like:

```
font family not found in Windows font database
```

To avoid this, switch to a modern graphics backend that uses
[systemfonts](https://CRAN.R-project.org/package=systemfonts) for font
resolution:

```r
# For knitr / R Markdown --- add to your setup chunk:
knitr::opts_chunk$set(dev = "ragg_png")

# For interactive use:
options(device = function(...) ragg::agg_png(tempfile(fileext = ".png"), ...))
```

Recommended backends:

| Backend               | Format | Package               |
|:----------------------|:-------|:----------------------|
| `ragg::agg_png()`     | PNG    | [ragg](https://CRAN.R-project.org/package=ragg)       |
| `svglite::svglite()`  | SVG    | [svglite](https://CRAN.R-project.org/package=svglite) |
| `grDevices::cairo_pdf()` | PDF | Base R (Cairo build)  |

Alternatively, use `render_mode = "path"` to bypass font lookup
entirely --- glyphs are drawn as vector paths, which works on all
devices but produces non-selectable text in PDF/SVG.

