---
title: "Introduction to gridmicrotex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gridmicrotex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300
)
library(gridmicrotex)
```

## What is gridmicrotex?

**gridmicrotex** renders LaTeX math equations as native R `grid` graphics
objects (grobs). It uses the [MicroTeX](https://github.com/NanoMichael/MicroTeX)
C++ library as its layout engine --- MicroTeX parses LaTeX, builds the TeX
box model, and computes exact glyph coordinates. The package intercepts
this layout data and maps it to native grid primitives (`pathGrob`,
`segmentsGrob`, `rectGrob`, `textGrob`), producing a `gTree` that works
on any R graphics device at any resolution.

**Key features:**

- No external LaTeX installation required --- MicroTeX is fully embedded
- Resolution-independent vector output on all R devices (PNG, PDF, SVG, ...)
- Full math support: fractions, roots, integrals, matrices, Greek letters,
  accents, delimiters, and more
- Multiple math fonts (Latin Modern Math, XITS Math, TeX Gyre DejaVu Math)
- Color support via `\textcolor{}`
- ggplot2 integration with `geom_latex()` and `element_latex()`
- CJK and multilingual text in `\text{}` blocks

## Basic usage

The core function is `latex_grob()`, which returns a grid grob:

```{r basic, fig.height=0.8, fig.width=2, out.width="50%"}
g <- latex_grob("\\frac{-b \\pm \\sqrt{b^2 - 4ac}}{2a}", fontsize = 24)
grid::grid.newpage()
grid::grid.draw(g)
```

For quick rendering, use `grid.latex()`:

```{r quick, fig.height=0.8, fig.width=2, out.width="50%"}
grid::grid.newpage()
grid.latex("\\sum_{i=1}^{n} x_i^2", fontsize = 28)
```

## Positioning and justification

Control placement with `x`, `y`, `hjust`, and `vjust`:

```{r positioning, fig.height=1.2, fig.width=2, out.width="50%"}
grid::grid.newpage()
grid.latex("E = mc^2", x = 0.2, y = 0.7, hjust = 0, fontsize = 24)
grid.latex("F = ma", x = 0.2, y = 0.3, hjust = 0, fontsize = 24)
```

## Colors

Set the formula color via `gp`, or use `\textcolor{}` within the LaTeX:

```{r colors, fig.height=0.6, fig.width=2, out.width="40%"}
grid::grid.newpage()
grid.latex(
  "\\textcolor{red}{\\alpha} + \\textcolor{blue}{\\beta} = \\gamma",
  fontsize = 28
)
```

## Math fonts

The package bundles Latin Modern Math (default), XITS Math, and
TeX Gyre DejaVu Math.
For most users, the easiest workflow is:

1. List available math fonts with `available_math_fonts()`
2. Select one with `set_math_font()`
3. Render formulas normally (no OTF/CLM paths needed)

```{r fonts-list}
available_math_fonts()
```

```{r fonts-default, fig.height=0.8, fig.width=2, out.width="50%"}
set_math_font("xits")
grid::grid.newpage()
grid.latex("\\int_0^1 f(x)\\,dx", fontsize = 24)

# Switch back to Latin Modern Math
set_math_font("lm")
```

You can still override the font per call via `math_font`:

```{r fonts, fig.height=1.5, fig.width=2, out.width="50%"}
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(2, 1)))
grid::pushViewport(grid::viewport(layout.pos.row = 1))
grid.latex("\\int_0^1 f(x)\\,dx", fontsize = 24)
grid::upViewport()
grid::pushViewport(grid::viewport(layout.pos.row = 2))
grid.latex("\\int_0^1 f(x)\\,dx", fontsize = 24, math_font = "xits")
grid::upViewport(2)
```

Use `available_math_fonts()` to list loaded fonts and `check_fonts()`
for a diagnostic report.

### Advanced: loading custom fonts

Use `load_font()` only when you need a custom font that is not already
bundled/loaded. In the current engine, custom loading still needs a
matching CLM metrics file (auto-discovered when possible):

```r
load_font("path/to/MyFont.otf")
```

You can find and download other fonts with CLM [here](https://github.com/NanoMichael/MicroTeX/tree/openmath/res)

### Render modes

gridmicrotex supports two rendering modes for math glyphs:
  
- **`"typeface"`** (default): Renders glyphs as native text using the
  math font's typeface. This produces selectable, searchable, and
  accessible text in PDF and SVG output. Requires the math font (e.g.,
  Latin Modern Math) to be installed on the system, and a device that
  supports font embedding (e.g., `ragg::agg_png()`,
  `svglite::svglite()`, `grDevices::cairo_pdf()`). On devices that
  do not support typeface rendering (e.g., the base `pdf()` device),
  the package automatically falls back to path mode with a warning.

- **`"path"`**: Renders each glyph as a filled vector path.
  This works on all R graphics devices and produces pixel-perfect output.
  However, text in PDF/SVG output is not selectable or searchable.

```r
# Default typeface mode (selectable text in PDF/SVG)
grid.latex("E = mc^2", fontsize = 24)

# Explicit path mode (works everywhere, but text is not selectable)
grid.latex("E = mc^2", fontsize = 24, render_mode = "path")
```

> **Important: Do not use `showtext::showtext_auto()` with typeface mode.**
> The [showtext](https://CRAN.R-project.org/package=showtext) package
> globally intercepts all text rendering and converts it to vector paths.
> This silently defeats typeface mode, causing all math glyphs to appear
> as paths instead of native text — even on devices like `svglite` and
> `ragg` that fully support font embedding. If you need showtext for
> other parts of your plot, disable it before drawing LaTeX formulas:
>
> ```r
> showtext::showtext_auto(FALSE)
> grid.latex("E = mc^2", fontsize = 24)  # typeface mode works correctly
> ```

## Querying dimensions

`latex_dims()` returns the bounding box of an expression:

```{r dims}
dims <- latex_dims("\\frac{a}{b}", fontsize = 20)
dims
```

This is useful for layout calculations and ensuring labels fit.

## Text rendering and CJK support

Non-math text inside `\text{}` is rendered as a standard `textGrob`.
Characters not in the math font (such as CJK) are rendered using the
font specified in `gp$fontfamily`:

```{r cjk, fig.height = 0.8, fig.width = 2.5, out.width="50%"}
grid::grid.newpage()
grid.latex("x^2 + \\text{你好}", fontsize = 24, gp = grid::gpar(fontfamily = "sans"))
```

## Supported LaTeX

gridmicrotex uses the MicroTeX engine, which is a **math formula renderer**,
not a full document typesetter. It covers the vast majority of math notation
you would use in plots and figures, but does not attempt to replace a full
LaTeX installation.

### Complicated examples

```{r complex-formula, fig.height = 3, fig.width=6, out.width="60%"}
grid::grid.newpage()
grid.latex(paste0(
      "\\begin{array}{l}",
      "  \\forall\\varepsilon\\in\\mathbb{R}_+^*\\ \\exists\\eta>0",
      "\\ |x-x_0|\\leq\\eta\\Longrightarrow|f(x)-f(x_0)|\\leq\\varepsilon\\\\",
      "  \\det",
      "  \\begin{bmatrix}",
      "      a_{11}&a_{12}&\\cdots&a_{1n}\\\\",
      "      a_{21}&\\ddots&&\\vdots\\\\",
      "      \\vdots&&\\ddots&\\vdots\\\\",
      "      a_{n1}&\\cdots&\\cdots&a_{nn}",
      "  \\end{bmatrix}",
      "  \\overset{\\mathrm{def}}{=}\\sum_{\\sigma\\in\\mathfrak{S}_n}",
      "\\varepsilon(\\sigma)\\prod_{k=1}^n a_{k\\sigma(k)}\\\\",
      "  \\int_0^\\infty{x^{2n} e^{-a x^2}\\,dx} = \\frac{2n-1}{2a}",
      " \\int_0^\\infty{x^{2(n-1)} e^{-a x^2}\\,dx}",
      " = \\frac{(2n-1)!!}{2^{n+1}} \\sqrt{\\frac{\\pi}{a^{2n+1}}}\\\\",
      "\\end{array}"
), fontsize = 16)
```


```{r table-multicolumn, fig.height = 3, fig.width=8, out.width="60%"}
grid::grid.newpage()

grid.latex(
  "
  \\newcolumntype{s}{>{\\color{#1234B6}}c}
\\begin{array}{|c|c|c|s|}
  \\hline
  \\rowcolor{Tan}\\multicolumn{4}{|c|}{\\textcolor{white}{\\bold{\\text{Table Head}}}}\\\\
  \\hline
  \\text{Matrix}&\\multicolumn{2}{|c|}{\\text{Multicolumns}}&\\text{Font size commands}\\\\
  \\hline
  \\begin{pmatrix}
      \\alpha_{11}&\\cdots&\\alpha_{1n}\\\\
      \\hdotsfor{3}\\\\
      \\alpha_{n1}&\\cdots&\\alpha_{nn}
  \\end{pmatrix}
  &\\large \\text{Left}&\\cellcolor{#00bde5}\\small \\textcolor{white}{\\text{\\bold{Right}}}
  &\\small \\text{small Small}\\\\
  \\hline
  \\multicolumn{4}{|c|}{\\text{Table Foot}}\\\\
  \\hline
\\end{array}
  ",
  fontsize = 22
)
```

```{r complicated-equation, fig.height = 9, fig.width=9, out.width="60%"}
grid::grid.newpage()
grid.latex(
  "\\definecolor{gris}{gray}{0.9}
\\definecolor{noir}{rgb}{0,0,0}
\\definecolor{bleu}{rgb}{0,0,1}
\\fatalIfCmdConflict{false}
\\newcommand{\\pa}{\\left|}
\\begin{array}{c}
  \\LaTeX\\\\
  \\begin{split}
      |I_2| &= \\pa\\int_0^T\\psi(t)\\left\\{ u(a,t)-\\int_{\\gamma(t)}^a \\frac{d\\theta}{k} (\\theta,t) \\int_a^\\theta c(\\xi)
          u_t (\\xi,t)\\,d\\xi\\right\\}dt\\right|\\\\
      &\\le C_6 \\Bigg|\\pa f \\int_\\Omega \\pa\\widetilde{S}^{-1,0}_{a,-}
          W_2(\\Omega, \\Gamma_1)\\right|\\ \\right|\\left| |u|\\overset{\\circ}{\\to} W_2^{\\widetilde{A}}(\\Omega\\Gamma_r,T)\\right|\\Bigg|\\\\
      &\\\\
      &\\begin{pmatrix}
          \\alpha&\\beta&\\gamma&\\delta\\\\
          \\aleph&\\beth&\\gimel&\\daleth\\\\
          \\mathfrak{A}&\\mathfrak{B}&\\mathfrak{C}&\\mathfrak{D}\\\\
          \\boldsymbol{\\mathfrak{a}}&\\boldsymbol{\\mathfrak{b}}&\\boldsymbol{\\mathfrak{c}}&\\boldsymbol{\\mathfrak{d}}
      \\end{pmatrix}
      \\quad{(a+b)}^{\\frac{n}{2}}=\\sqrt{\\sum_{k=0}^n\\tbinom{n}{k}a^kb^{n-k}}\\quad
          \\Biggl(\\biggl(\\Bigl(\\bigl(()\\bigr)\\Bigr)\\biggr)\\Biggr)\\\\
      &\\forall\\varepsilon\\in\\mathbb{R}_+^*\\ \\exists\\eta>0\\ |x-x_0|\\leq\\eta\\Longrightarrow|f(x)-f(x_0)|\\leq\\varepsilon\\\\
      &\\det
      \\begin{bmatrix}
          a_{11}&a_{12}&\\cdots&a_{1n}\\\\
          a_{21}&\\ddots&&\\vdots\\\\
          \\vdots&&\\ddots&\\vdots\\\\
          a_{n1}&\\cdots&\\cdots&a_{nn}
      \\end{bmatrix}
      \\overset{\\mathrm{def}}{=}\\sum_{\\sigma\\in\\mathfrak{S}_n}\\varepsilon(\\sigma)\\prod_{k=1}^n a_{k\\sigma(k)}\\\\
      &\\Delta f(x,y)=\\frac{\\partial^2f}{\\partial x^2}+\\frac{\\partial^2f}{\\partial y^2}\\qquad\\qquad \\fcolorbox{noir}{gris}
          {n!\\underset{n\\rightarrow+\\infty}{\\sim} {\\left(\\frac{n}{e}\\right)}^n\\sqrt{2\\pi n}}\\\\
      &\\sideset{_\\alpha^\\beta}{_\\gamma^\\delta}{
      \\begin{pmatrix}
          a&b\\\\
          c&d
      \\end{pmatrix}}
      \\xrightarrow[T]{n\\pm i-j}\\sideset{^t}{}A\\xleftarrow{\\overrightarrow{u}\\wedge\\overrightarrow{v}}
          \\underleftrightarrow{\\iint_{\\mathds{R}^2}e^{-\\left(x^2+y^2\\right)}\\,\\mathrm{d}x\\mathrm{d}y}
  \\end{split}\\\\
  \\rotatebox{30}{\\sum_{n=1}^{+\\infty}}\\quad\\mbox{Mirror rorriM}\\reflectbox{\\mbox{Mirror rorriM}}
\\end{array}",
  fontsize = 22
)
```



### What is not supported

MicroTeX is a **math formula renderer**, not a full LaTeX engine. The
following are outside its scope:

- **Document structure**: `\section`, `\begin{document}`, page layout,
  headers/footers, `\tableofcontents`
- **Package loading**: `\usepackage{}` --- all supported commands are
  built in
- **Paragraph text**: line breaking, hyphenation, justified paragraphs
- **TikZ / PGF** drawing commands
- **Images**: `\includegraphics`
- **Cross-references**: `\label`, `\ref`, `\cite`, bibliographies
- **Theorem environments**: `\begin{theorem}`, `\begin{proof}`
- **Lists**: `itemize`, `enumerate`, `description`
- **Some amsmath commands**: `\substack`, `\tag`, equation numbering

For most statistical graphics use cases --- axis labels, annotations,
legends, and in-plot formulas --- the supported feature set is more
than sufficient.

## Comparison with alternatives

| Approach       | LaTeX required? | Device independent? | Vector? | Math coverage |
|:---------------|:---------------:|:-------------------:|:-------:|:-------------:|
| `tikzDevice`   | Yes             | No                  | Yes     | Full          |
| `xdvir`        | Yes             | No                  | Yes     | Full          |
| `latexpdf`     | Yes             | No                  | Yes     | Full (tables) |
| `latex2exp`    | No              | Yes                 | Yes     | Limited       |
| `plotmath`     | No              | Yes                 | Yes     | Limited       |
| **gridmicrotex** | **No**       | **Yes**             | **Yes** | **Broad**     |


## Graphics backend

The default graphics device on Windows (`windows()`) and macOS
(`quartz()`) may not find the bundled math fonts, producing warnings
like:

```
font family not found in Windows font database
```

To avoid this, switch to a modern graphics backend that uses
[systemfonts](https://CRAN.R-project.org/package=systemfonts) for font
resolution:

```r
# For knitr / R Markdown --- add to your setup chunk:
knitr::opts_chunk$set(dev = "ragg_png")

# For interactive use:
options(device = function(...) ragg::agg_png(tempfile(fileext = ".png"), ...))
```

Recommended backends:

| Backend               | Format | Package               |
|:----------------------|:-------|:----------------------|
| `ragg::agg_png()`     | PNG    | [ragg](https://CRAN.R-project.org/package=ragg)       |
| `svglite::svglite()`  | SVG    | [svglite](https://CRAN.R-project.org/package=svglite) |
| `grDevices::cairo_pdf()` | PDF | Base R (Cairo build)  |

Alternatively, use `render_mode = "path"` to bypass font lookup
entirely --- glyphs are drawn as vector paths, which works on all
devices but produces non-selectable text in PDF/SVG.

