---
title: "Rendering Markdown with Math"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Markdown with Math}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "ragg_png"
)
library(gridmicrotex)
library(grid)
```

Plot labels are rarely pure math. A title might be a bold phrase with a
symbol in it; a caption might be a short paragraph that happens to
contain an equation. `gridmicrotex` renders
[CommonMark](https://commonmark.org) markdown *and* LaTeX math together,
in one grob, with no external LaTeX installation.

It does this by translating the markdown to LaTeX and handing that to
MicroTeX --- the same engine `latex_grob()` uses. The `$...$` math in
your markdown is passed straight through untouched, so anything MicroTeX
can typeset works inside a markdown label.

There are two entry points:

- **`markdown_grob()`** (and `grid.markdown()`) for a single flowing
  run of text --- the right choice for a title or an axis label.
- **`markdown_box_grob()`** for a whole block document --- headings,
  lists, block quotes, code, tables --- inside an optional padded,
  filled box.

## Inline markdown

`markdown_grob()` handles the inline part of markdown: `**bold**`,
`*italic*`, `` `code` ``, `~~strikethrough~~`, and any `$math$`.

```{r inline, fig.height = 0.7, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  "The **fitted** slope is $\\beta_1 = 0.42$ (*p* < 0.001).",
  x = 0.02, hjust = 0, gp = gpar(fontsize = 20)
)
```

Because the math is real MicroTeX, the full engine is available inside a
markdown string --- fractions, sums, matrices, Greek, accents:

```{r inline-math, fig.height = 0.9, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  paste(
    "The estimator $\\hat{\\beta} = (X^\\top X)^{-1} X^\\top y$",
    "has variance $\\sigma^2$."
  ),
  x = 0.02, hjust = 0, gp = gpar(fontsize = 18)
)
```

### Wrapping long labels

Set `max_width` (in big points) to wrap a long label over several lines.

```{r wrap, fig.height = 1.4, fig.width = 4, out.width = "60%"}
grid.newpage()
grid.markdown(
  paste(
    "This is a **long caption** with some $x^2$ math that is wide enough",
    "to wrap across several lines when a maximum width is set."
  ),
  x = 0.02, y = 0.95, hjust = 0, vjust = 1,
  max_width = 4 * 72,          # 4 inches, in big points
  gp = gpar(fontsize = 14)
)
```

### Colour and inline HTML

Markdown has no syntax for colour, underline, super/subscript, highlight
or size --- CommonMark and GFM simply do not define one. The conformant
way to ask for them is the inline HTML that CommonMark already includes,
so that is what `gridmicrotex` reads. Each tag does what HTML's own
default rendering says it should:

| tag | effect |
|---|---|
| `<b>`, `<strong>` | bold |
| `<i>`, `<em>`, `<cite>`, `<dfn>`, `<var>` | italic |
| `<code>`, `<kbd>`, `<samp>`, `<tt>` | monospace |
| `<u>`, `<ins>` | underline |
| `<s>`, `<del>`, `<strike>` | strikethrough |
| `<sub>`, `<sup>` | sub / superscript |
| `<mark>` | yellow highlight |
| `<small>`, `<big>` | smaller / larger |
| `<q>` | wrapped in quotation marks |
| `<br>` | line break |
| `<span style="…">` | `color`, `text-decoration`, `font-size`, `font-family` |

In a `style` attribute: `color` takes any R colour name, the nine CSS
names R happens to lack (`crimson`, `teal`, `rebeccapurple` and friends),
`#rgb`, `#rrggbb` or `rgb()` --- but note that `green`, `gray`, `grey`,
`maroon` and `purple` keep their R values rather than their (darker) CSS
ones, so that they mean the same thing here as in `gpar(col = ...)`;
`text-decoration` takes `underline` or
`line-through`; `font-size` takes `pt`, `px`, `in`, `cm`, `mm`, `em`,
`rem`, `%`, `smaller` or `larger`; `font-family` takes the CSS generics
`monospace`, `sans-serif` and `serif`, or any font name, including via a
fallback list like `'Courier New', monospace`. Any other property is
ignored.

```{r html-css, fig.height = 0.7, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  paste0(
    '<span style="font-size:26pt;color:#2E5E8E">Big</span> and ',
    '<span style="font-family:serif">serif</span> and ',
    '<span style="font-family:monospace">mono</span>.'
  ),
  x = 0.02, hjust = 0, gp = gpar(fontsize = 16)
)
```

### Which fonts are available

A family name goes straight to `gp$fontfamily`, so **the device resolves
it**, exactly as for any other grid text:

| device | sees |
|---|---|
| `ragg`, `svglite` | any installed family, plus `systemfonts::register_font()` |
| cairo (`png(type = "cairo")`, `cairo_pdf()`) | installed families |
| `windows()`, `quartz()` | the OS font list |
| base `pdf()`, `postscript()` | **only** what `pdfFonts()` declares --- a named family will not resolve |

An unavailable font falls back silently, as it does for
`gpar(fontfamily = ...)`. To use a font file that is not installed
system-wide, register it first:

```r
systemfonts::register_font(name = "MyFont", plain = "path/to/MyFont.otf")
grid.markdown('<span style="font-family:MyFont">x</span>')
```

`load_font()` is *not* the function for this --- it registers *math*
fonts with MicroTeX, a different mechanism.

A span's own family wins over `gp$fontfamily`. One thing it does not
override: the width of the spaces *between* its words, which still comes
from `gp$fontfamily`. Set both to the same family if that shows.

```{r html-spans, fig.height = 0.7, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  paste0(
    'The <span style="color:#B22222">**residual**</span> for ',
    "H<sub>0</sub> is <u>within</u> $2\\sigma$."
  ),
  x = 0.02, hjust = 0, gp = gpar(fontsize = 20)
)
```

Tags nest, and markdown and `$math$` keep working inside them, so
`<u>**bold underlined**</u>` and `**<u>the same thing</u>**` render
alike. Everything else is dropped and the text inside it kept --- which
is also what a browser does with `<a>`, `<abbr>` or a bare `<span>`,
since those have no default rendering of their own.

Two `<br>` in a row leave a blank line, as in HTML:

```{r html-br, fig.height = 1.1, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  "**Model fit**<br><br>Deviance <mark>412.7</mark> on term<sub>1</sub>.",
  x = 0.02, y = 0.85, vjust = 1, hjust = 0, gp = gpar(fontsize = 18)
)
```

## Block documents

`markdown_box_grob()` lays out a full markdown document. Each block ---
heading, paragraph, list, quote, code, table --- becomes its own grob,
stacked top to bottom. The `width` you give is the box width; prose
wraps to fit it.

```{r block, fig.height = 4.4, fig.width = 4.6, out.width = "80%"}
md <- "
# Model summary

The slope is $\\beta_1$ with *p* < 0.001, and this paragraph is long
enough that it wraps inside the column.

## Diagnostics

- residuals look **fine**
- $R^2 = 0.87$
- no influential points

> Assumptions were checked and hold.
"

grid.newpage()
grid.draw(markdown_box_grob(
  md,
  width   = unit(4.6, "in"),
  padding = unit(10, "pt"),
  box_gp  = gpar(fill = "grey97", col = "grey40"),
  gp      = gpar(fontsize = 13)
))
```

Set `box_gp = NULL` to draw no box, or give the `r` argument a value
such as `unit(6, "pt")` for rounded corners. `padding` and `margin`
accept a single unit or a length-4 unit giving top, right, bottom, left.

### Lists, tasks, and tables

Ordered and unordered lists, GFM task lists, and tables all render.
Markdown tables become real typeset tables, with column alignment taken
from the `|:---:|` markers --- something a text renderer cannot do.

```{r lists-tables, fig.height = 3.6, fig.width = 4.4, out.width = "80%"}
md <- "
### Checklist

- [x] fit the model
- [ ] write it up

Coefficients:

| term | estimate | *p* |
|:-----|---------:|:---:|
| intercept | 30.1 | *** |
| slope | -5.3 | *** |
"

grid.newpage()
grid.draw(markdown_box_grob(
  md,
  width   = unit(4.4, "in"),
  padding = unit(8, "pt"),
  gp      = gpar(fontsize = 13)
))
```

### Images

An image on its own line is drawn as a raster, scaled to the column
width. `png` and `jpeg` are optional --- if neither is installed, or the
file is missing, the image degrades to its alt text.

```{r image, fig.height = 2.4, fig.width = 4, out.width = "70%"}
# Make a small PNG to embed (any .png / .jpeg path works).
img <- tempfile(fileext = ".png")
ragg::agg_png(img, width = 300, height = 90, res = 100)
grid.rect(gp = gpar(fill = "steelblue", col = NA))
grid.text("a figure", gp = gpar(col = "white", fontsize = 18))
invisible(dev.off())

md <- sprintf("
A figure with a caption below it.

![a blue box](%s)

The caption explains the figure.
", img)

grid.newpage()
grid.draw(markdown_box_grob(
  md, width = unit(4, "in"), padding = unit(8, "pt"),
  gp = gpar(fontsize = 13)
))
```

## Line breaking and justification

Two optional refinements apply once text wraps (both need `max_width`,
and both are off by default so the output matches R's usual ragged-right
text).

`justify = TRUE` stretches the interword spaces so every line but the
last fills the width exactly. `line_break = "optimal"` chooses the line
breaks by total fit rather than greedily, so the paragraph reads more
evenly. Note that `gridmicrotex` does not hyphenate, so a justified
*narrow* column can open wide word gaps.

```{r justify, fig.height = 2.6, fig.width = 4, out.width = "70%"}
prose <- paste(rep(
  "The quick brown fox jumps over the lazy dog.", 3), collapse = " ")

grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 1)))

pushViewport(viewport(layout.pos.row = 1))
grid.text("ragged (default)", x = 0.02, y = 0.95, hjust = 0, vjust = 1,
          gp = gpar(fontsize = 8, col = "grey40"))
grid.markdown(prose, x = 0.02, y = 0.75, hjust = 0, vjust = 1,
              max_width = 3.6 * 72, gp = gpar(fontsize = 12))
popViewport()

pushViewport(viewport(layout.pos.row = 2))
grid.text("justified + optimal", x = 0.02, y = 0.95, hjust = 0, vjust = 1,
          gp = gpar(fontsize = 8, col = "grey40"))
grid.markdown(prose, x = 0.02, y = 0.75, hjust = 0, vjust = 1,
              max_width = 3.6 * 72, justify = TRUE, line_break = "optimal",
              gp = gpar(fontsize = 12))
popViewport()
popViewport()
```

## In ggplot2

Markdown works as a geom and as a theme element, mirroring
`geom_latex()` and `element_latex()` --- see the *ggplot2 integration*
vignette. In short:

```{r ggplot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.height = 3, fig.width = 5, out.width = "80%"}
library(ggplot2)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(
    title = "*Fitted* model: $\\hat{y} = \\beta_0 + \\beta_1 x$",
    x     = "**weight** ($10^3$ lbs)",
    y     = "*efficiency* $\\eta$"
  ) +
  theme(
    plot.title   = element_markdown(fontsize = 14),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
```

## What is and isn't supported

The parser is the full CommonMark specification plus the GitHub
extensions (tables, strikethrough, task lists), so parsing is never the
limitation. On the rendering side:

- **Inline images** keep their alt text only. A block-level image (alone
  on its line) is drawn; one in the middle of a sentence is not, because
  a single MicroTeX line has no raster.
- **Links** keep their text; the destination is dropped, since a grob
  cannot be a hyperlink.
- **Block-level HTML** (a `<div>`, say) is dropped rather than typeset.
  Inline HTML is partly rendered --- see below.
- **Syntax highlighting** of code blocks is not applied --- code is set
  in a monospace font, unhighlighted.
- **Monospace spacing** depends on the device. `ragg`, `pdf()` and
  `cairo_pdf()` report correct metrics for the `"mono"` family; the base
  `png(type = "cairo")` and `svglite` report the *sans* widths for it, so
  code set with them comes out slightly tight. Use `ragg::agg_png()` if
  that matters.

Everything else --- emphasis, headings, lists, quotes, tables, rules,
block images, and of course `$math$` --- renders.
