---
title: "Rendering Markdown with Math"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Markdown with Math}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "ragg_png"
)
library(gridmicrotex)
library(grid)
```

Plot labels are rarely pure math. A title might be a bold phrase with a
symbol in it; a caption might be a short paragraph that happens to
contain an equation. `gridmicrotex` renders
[CommonMark](https://commonmark.org) markdown *and* LaTeX math together,
in one grob, with no external LaTeX installation.

It does this by translating the markdown to LaTeX and handing that to
MicroTeX --- the same engine `latex_grob()` uses. The `$...$` math in
your markdown is passed straight through untouched, so anything MicroTeX
can typeset works inside a markdown label.

There are two entry points:

- **`markdown_grob()`** (and `grid.markdown()`) for a single flowing
  run of text --- the right choice for a title or an axis label.
- **`markdown_box_grob()`** for a whole block document --- headings,
  lists, block quotes, code, tables --- inside an optional padded,
  filled box.

## Inline markdown

`markdown_grob()` handles the inline part of markdown: `**bold**`,
`*italic*`, `` `code` ``, `~~strikethrough~~`, and any `$math$`.

```{r inline, fig.height = 0.7, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  "The **fitted** slope is $\\beta_1 = 0.42$ (*p* < 0.001).",
  x = 0.02, hjust = 0, gp = gpar(fontsize = 20)
)
```

Because the math is real MicroTeX, the full engine is available inside a
markdown string --- fractions, sums, matrices, Greek, accents:

```{r inline-math, fig.height = 0.9, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  paste(
    "The estimator $\\hat{\\beta} = (X^\\top X)^{-1} X^\\top y$",
    "has variance $\\sigma^2$."
  ),
  x = 0.02, hjust = 0, gp = gpar(fontsize = 18)
)
```

### Wrapping long labels

Set `max_width` (in big points) to wrap a long label over several lines.

```{r wrap, fig.height = 1.4, fig.width = 4, out.width = "60%"}
grid.newpage()
grid.markdown(
  paste(
    "This is a **long caption** with some $x^2$ math that is wide enough",
    "to wrap across several lines when a maximum width is set."
  ),
  x = 0.02, y = 0.95, hjust = 0, vjust = 1,
  max_width = 4 * 72,          # 4 inches, in big points
  gp = gpar(fontsize = 14)
)
```

## Block documents

`markdown_box_grob()` lays out a full markdown document. Each block ---
heading, paragraph, list, quote, code, table --- becomes its own grob,
stacked top to bottom. The `width` you give is the box width; prose
wraps to fit it.

```{r block, fig.height = 4.4, fig.width = 4.6, out.width = "80%"}
md <- "
# Model summary

The slope is $\\beta_1$ with *p* < 0.001, and this paragraph is long
enough that it wraps inside the column.

## Diagnostics

- residuals look **fine**
- $R^2 = 0.87$
- no influential points

> Assumptions were checked and hold.
"

grid.newpage()
grid.draw(markdown_box_grob(
  md,
  width   = unit(4.6, "in"),
  padding = unit(10, "pt"),
  box_gp  = gpar(fill = "grey97", col = "grey40"),
  gp      = gpar(fontsize = 13)
))
```

Set `box_gp = NULL` to draw no box, or give the `r` argument a value
such as `unit(6, "pt")` for rounded corners. `padding` and `margin`
accept a single unit or a length-4 unit giving top, right, bottom, left.

### Lists, tasks, and tables

Ordered and unordered lists, GFM task lists, and tables all render.
Markdown tables become real typeset tables, with column alignment taken
from the `|:---:|` markers --- something a text renderer cannot do.

```{r lists-tables, fig.height = 3.6, fig.width = 4.4, out.width = "80%"}
md <- "
### Checklist

- [x] fit the model
- [ ] write it up

Coefficients:

| term | estimate | *p* |
|:-----|---------:|:---:|
| intercept | 30.1 | *** |
| slope | -5.3 | *** |
"

grid.newpage()
grid.draw(markdown_box_grob(
  md,
  width   = unit(4.4, "in"),
  padding = unit(8, "pt"),
  gp      = gpar(fontsize = 13)
))
```

### Images

An image on its own line is drawn as a raster, scaled to the column
width. `png` and `jpeg` are optional --- if neither is installed, or the
file is missing, the image degrades to its alt text.

```{r image, fig.height = 2.4, fig.width = 4, out.width = "70%"}
# Make a small PNG to embed (any .png / .jpeg path works).
img <- tempfile(fileext = ".png")
ragg::agg_png(img, width = 300, height = 90, res = 100)
grid.rect(gp = gpar(fill = "steelblue", col = NA))
grid.text("a figure", gp = gpar(col = "white", fontsize = 18))
invisible(dev.off())

md <- sprintf("
A figure with a caption below it.

![a blue box](%s)

The caption explains the figure.
", img)

grid.newpage()
grid.draw(markdown_box_grob(
  md, width = unit(4, "in"), padding = unit(8, "pt"),
  gp = gpar(fontsize = 13)
))
```

## Line breaking and justification

Two optional refinements apply once text wraps (both need `max_width`,
and both are off by default so the output matches R's usual ragged-right
text).

`justify = TRUE` stretches the interword spaces so every line but the
last fills the width exactly. `line_break = "optimal"` chooses the line
breaks by total fit rather than greedily, so the paragraph reads more
evenly. Note that `gridmicrotex` does not hyphenate, so a justified
*narrow* column can open wide word gaps.

```{r justify, fig.height = 2.6, fig.width = 4, out.width = "70%"}
prose <- paste(rep(
  "The quick brown fox jumps over the lazy dog.", 3), collapse = " ")

grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 1)))

pushViewport(viewport(layout.pos.row = 1))
grid.text("ragged (default)", x = 0.02, y = 0.95, hjust = 0, vjust = 1,
          gp = gpar(fontsize = 8, col = "grey40"))
grid.markdown(prose, x = 0.02, y = 0.75, hjust = 0, vjust = 1,
              max_width = 3.6 * 72, gp = gpar(fontsize = 12))
popViewport()

pushViewport(viewport(layout.pos.row = 2))
grid.text("justified + optimal", x = 0.02, y = 0.95, hjust = 0, vjust = 1,
          gp = gpar(fontsize = 8, col = "grey40"))
grid.markdown(prose, x = 0.02, y = 0.75, hjust = 0, vjust = 1,
              max_width = 3.6 * 72, justify = TRUE, line_break = "optimal",
              gp = gpar(fontsize = 12))
popViewport()
popViewport()
```

## In ggplot2

Markdown works as a geom and as a theme element, mirroring
`geom_latex()` and `element_latex()` --- see the *ggplot2 integration*
vignette. In short:

```{r ggplot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.height = 3, fig.width = 5, out.width = "80%"}
library(ggplot2)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(
    title = "*Fitted* model: $\\hat{y} = \\beta_0 + \\beta_1 x$",
    x     = "**weight** ($10^3$ lbs)",
    y     = "*efficiency* $\\eta$"
  ) +
  theme(
    plot.title   = element_markdown(fontsize = 14),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
```

## What is and isn't supported

The parser is the full CommonMark specification plus the GitHub
extensions (tables, strikethrough, task lists), so parsing is never the
limitation. On the rendering side:

- **Inline images** keep their alt text only. A block-level image (alone
  on its line) is drawn; one in the middle of a sentence is not, because
  a single MicroTeX line has no raster.
- **Links** keep their text; the destination is dropped, since a grob
  cannot be a hyperlink.
- **Raw HTML** is dropped rather than typeset.
- **Syntax highlighting** of code blocks is not applied --- code is set
  in a monospace font, unhighlighted.

Everything else --- emphasis, headings, lists, quotes, tables, rules,
block images, and of course `$math$` --- renders.
