---
title: "Rendering Markdown with Math"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Markdown with Math}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  dev = "png"
)
library(gridmicrotex)
library(grid)
```

gridmicrotex renders [CommonMark](https://commonmark.org) markdown with
LaTeX math between `$...$`. There are two functions:

- `markdown_grob()` (and `grid.markdown()`) for a single label, such as
  a title.
- `markdown_box_grob()` for a document with headings, lists, quotes,
  code and tables.

Fonts, options and LaTeX support are the same as for `latex_grob()`
(see `vignette("getting-started")`).

## Inline markdown

```{r inline, fig.height = 0.8, fig.width = 6, out.width = "90%"}
grid.newpage()
grid.markdown(
  r"(The **fitted** $\hat{\beta} = (X^\top X)^{-1} X^\top y$ has ~~no~~ `se` = *0.42*.)",
  x = 0.02, hjust = 0, gp = gpar(fontsize = 13)
)
```

### Inline HTML

For colour, underline, sub- and superscripts, highlight and size, use
inline HTML:

| tag | effect |
|---|---|
| `<b>`, `<strong>` | bold |
| `<i>`, `<em>`, `<cite>`, `<dfn>`, `<var>` | italic |
| `<code>`, `<kbd>`, `<samp>`, `<tt>` | monospace |
| `<u>`, `<ins>` | underline |
| `<s>`, `<del>`, `<strike>` | strikethrough |
| `<sub>`, `<sup>` | sub / superscript |
| `<mark>` | yellow highlight |
| `<small>`, `<big>` | smaller / larger |
| `<q>` | quotation marks |
| `<ruby>`, `<rt>` | annotation above the text (furigana) |
| `&nbsp;` | non-breaking space |
| `<br>` | line break |
| `<span style="…">` | `color`, `font-size`, `font-family`, `text-decoration`, and the rest of `?md_style` |

```{r html-css, fig.height = 1.3, fig.width = 6, out.width = "90%", dev = "ragg_png", dev.args = list()}
grid.newpage()
grid.markdown(
  r"(<span style="font-size:22pt;color:#2E5E8E">Model fit</span>,
     <span style="font-family:serif">serif</span> and
     <span style="font-family:monospace">mono</span>.<br><br>
     The <span style="color:#B22222">**residual**</span> for H<sub>0</sub>
     is <u>within</u> <mark>$2\sigma$</mark>.)",
  x = 0.02, y = 0.9, hjust = 0, vjust = 1, gp = gpar(fontsize = 13)
)
```

Tags nest, and markdown and math work inside them. Other tags are
dropped and their text kept.

## Documents

`markdown_box_grob()` stacks the blocks of a document inside an optional
box. Text wraps to `width`. Code below is quoted with `~~~r` and `~~~`, 
instead of `"```"` to avoid conflicts with R Markdown of this vignette.
But you should use `"```"` in your own code.

```{r block, fig.height = 3.6, fig.width = 4.6, out.width = "80%"}
md <- r"(
# Model summary

The slope is $\beta_1$ with *p* < 0.001, and this paragraph is long
enough that it wraps inside the column.

## Diagnostics

- residuals look **fine**
- $R^2 = 0.87$
- no influential points

~~~r
fit <- lm(y ~ x, data = d)  # refit
if (anyNA(d)) {
    stop("missing values")
}
~~~

> Assumptions were checked and hold.
)"

grid.newpage()
grid.draw(markdown_box_grob(
  md,
  width   = unit(4.6, "in"),
  padding = unit(10, "pt"),
  box_gp  = gpar(fill = "grey97", col = "grey40"),
  gp      = gpar(fontsize = 13),
  style = markdown_style(css = "pre { background: #ece2f0 }") # Code background
))
```

`box_gp = NULL` draws no box, and `r` rounds its corners. `padding` and
`margin` take one unit, or four for top, right, bottom and left.

### Code

A fenced code block that names its language is highlighted.
`available_highlighters()` lists the languages; `register_highlighter()`
adds one from a [KDE syntax file](https://kate-editor.org/syntax/).
Colours use the same class names as Pandoc and knitr, so a Pandoc theme
can be pasted in:

```r
markdown_style(css = ".co { color: #59636E } .kw { color: #CF222E }")
```

### Lists and tables

Task lists and tables work, with column alignment from the `|:---:|`
markers:

```{r lists-tables, fig.height = 2, fig.width = 4.4, out.width = "80%"}
md <- r"(
### Checklist

- [x] fit the model
- [ ] write it up

Coefficients:

| term | $\beta$  | *p* |
|:-----|------------:|:---:|
| intercept | 30.1 | *** |
| slope | -5.3 | *** |
)"

grid.newpage()
grid.draw(markdown_box_grob(
  md,
  width   = unit(4.4, "in"),
  padding = unit(8, "pt"),
  gp      = gpar(fontsize = 13)
))
```

### Images

An image on its own line is a block, scaled to fit; one inside a
sentence is inline. Images must be local PNG, JPEG or SVG files.

```{r image, fig.height = 1.8, fig.width = 4, out.width = "70%"}
img <- system.file("img", "Rlogo.png", package = "png")

md <- sprintf("
A figure with a caption below it.

![the R logo](%s)

The caption explains the figure.
", img)

grid.newpage()
grid.draw(markdown_box_grob(
  md, width = unit(4, "in"), padding = unit(8, "pt"),
  gp = gpar(fontsize = 13)
))
```

## Styling

`style` takes CSS, as text or a `.css` file:

```{r style-css, fig.height = 1.2, fig.width = 5, out.width = "85%"}
doc <- r"(
# Results

The slope is $\beta_1$ with *p* < 0.001.

> Worth a second look.
)"

grid.newpage()
grid.draw(markdown_box_grob(
  doc,
  width = unit(4.5, "in"), padding = unit(10, "pt"),
  style = "
    h1         { color: steelblue; font-size: 1.8rem }
    blockquote { color: grey40; border-left: 3px solid steelblue }
  ",
  gp = gpar(fontsize = 13)
))
```

Or the same in R:

```{r style-r, eval = FALSE}
markdown_style(
  h1         = md_style(color = "steelblue", font_size = 1.8),
  blockquote = md_style(color = "grey40",
                        border_left = "3px solid steelblue")
)
```

A bare number is a multiple of the body font size (`rem`). Tags are named
as in HTML (`p`, `h1`, `li`, `blockquote`, `pre`, `code`, `table`, ...).
`?md_style` lists the supported properties; others are ignored.

`markdown_style("github")` starts from a GitHub-like preset, and
`latex_options(markdown_style = )` sets a default for the session.

### Styling one part

Wrap a part in a `<div>` with a `class` or `style`, leaving a blank line
after the opening tag and before the closing one:

```{r style-div, fig.height = 1, fig.width = 5, out.width = "85%"}
doc <- '
Ordinary text.

<div class="note">

**Note.** This chunk is indented and set apart.

</div>

Ordinary text again.
'

grid.newpage()
grid.draw(markdown_box_grob(
  doc,
  width = unit(4.5, "in"), padding = unit(10, "pt"),
  style = ".note { color: grey35; padding-left: 1.5rem }",
  gp = gpar(fontsize = 13)
))
```

`<span class="...">` does the same inside a line.

### Styling tables

| CSS | effect |
|---|---|
| `table { border-color }` | colour of every rule |
| `tr { background }` | fills a row |
| `td`, `th ` `{ background }` | fills a cell |
| `tr { border-bottom }` | a rule under each row |
| `td { border-left }` | rules between columns |
| `td { padding-left }` | the gap between columns |
| `table { table-layout: fixed }` | columns of equal width |

A table that is too wide for the box wraps its widest columns.

```{r style-table, fig.height = 2, fig.width = 5, out.width = "85%"}
tbl <- r"(
| Term | Meaning |
|:-----|:--------|
| $\beta_1$ | the slope, which needs a good deal of room to explain |
| $\sigma$ | the residual standard deviation |
)"

grid.newpage()
grid.draw(markdown_box_grob(
  tbl,
  width = unit(4, "in"), padding = unit(8, "pt"),
  style = "
    table { table-layout: fixed; border-color: #D1D9E0 }
    th    { background: #F6F8FA }
    tr    { border-bottom: 1px solid #D1D9E0 }
  ",
  gp = gpar(fontsize = 13)
))
```

## LaTeX inside markdown

For what markdown cannot do, such as merged or coloured table cells,
write LaTeX in a `$$...$$` block:

```{r latex-escape, fig.height = 2.4, fig.width = 4.6, out.width = "80%"}
tbl <- r"($$\begin{array}{|l|c|c|}\hline
\rowcolor{#F6F8FA}\multicolumn{3}{|c|}{\textbf{Model comparison}}\\\hline
\textbf{Model}&\textbf{AIC}&\textbf{R}^2\\\hline
\text{linear}&214.3&0.71\\
\cellcolor{#DDF0DD}\text{quadratic}&\cellcolor{#DDF0DD}201.8&\cellcolor{#DDF0DD}0.87\\\hline
\end{array}$$)"

grid.newpage()
grid.draw(markdown_box_grob(
  paste0("Compare the two fits.\n\n", tbl, "\n\nThe quadratic wins."),
  width = unit(4.6, "in"), padding = unit(10, "pt"),
  style = "math { color: #1F3864; font-size: 1.1rem }",
  gp = gpar(fontsize = 13)
))
```

Inside `\textbf{}` you are in text, so write `\textbf{R}^2`, not
`\textbf{R^2}`. CSS table rules do not apply to a LaTeX table; style it
in LaTeX.

## Limitations

- Links show their text only.
- Footnotes are placed at the bottom in `markdown_box_grob()`; in
  `markdown_grob()` only the marker is shown.
- HTML blocks other than `<div>` and `<img>` are dropped, `<table>`
  included. Use markdown tables.
- Markdown tables cannot merge cells; use LaTeX (above).
- Small caps are not supported.
- The syntax class names (`co`, `st`, `kw`, `dt`, ...) also apply to your
  own `<div class>`; choose other names for your classes.
- Code is a little narrow on `png(type = "cairo")` and `svglite`; use
  `ragg::agg_png()` if it matters.
